---
title: "Ordination"
output: html_notebook
author: Gati Aher
date: 02/12/2022
---

Goal: explore sample gradient patterns and identify strong clusters

References
General
* Follow Professor Nicholas Ollberding's [Introduction to the Statistical Analysis of Microbiome Data in R](https://www.nicholas-ollberding.com/post/introduction-to-the-statistical-analysis-of-microbiome-data-in-r/)

Phyloseq:
* [Importing phyloseq Data](https://joey711.github.io/phyloseq/import-data.html)
* [How to save the phyloseq object generated in dada2' tutorial for later use #107](https://github.com/benjjneb/dada2/issues/107)
* [Functions for Accessing and (Pre)Processing Dat](https://joey711.github.io/phyloseq/preprocess.html)
* [Bonus: Handoff to phyloseq](http://benjjneb.github.io/dada2/tutorial.html)

# Install Packages

```{r}
# .cran_packages <- c("tidyverse", "cowplot", "picante", "vegan", "HMP", "dendextend", "rms", "devtools")
# .inst <- .cran_packages %in% installed.packages()
# if(any(!.inst)) {
#   install.packages(.cran_packages[!.inst])
# }
```

```{r}
# .bioc_packages <- c("phyloseq", "Biostrings", "DESeq2", "microbiome", "metagenomeSeq", "ALDEx2")
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(.bioc_packages, version = "3.14")
```

# Loading Required Packages

```{r}
library("ggplot2"); packageVersion("ggplot2") # plotting
```
```{r}
library("tidyverse"); packageVersion("tidyverse") # util              
```
```{r}
library("dplyr"); packageVersion("dplyr") # util
```
```{r}
library("phyloseq"); packageVersion("phyloseq") # data object                 
```
```{r}
library("ape"); packageVersion("ape") # taxa string manipulation
```
```{r}
library(microbiome); packageVersion("microbiome") # clr transformation
```

```{r}
library(dendextend); packageVersion("dendextend") # extend customization of dendograms
```

```{r}
library(vegan); packageVersion("vegan") # bray-curtis and other metrics
```

---

# Initialization

```{r}
# ggplot2 package theme set
theme_set(theme_bw())

# control randomization
set.seed(42)

# save outputs
raw_location = "../data/raw/FCF/"
save_location = "../data/processed_FCF/"

# PARAMETERS
MIN_SAMPLE_RC = 5000 # only keep samples with at least 5k reads
MIN_PERCENT_IN_ANY_SAMPLE = 0.001 # only keep taxa that occurred at an abundance of greater than 0.1% in any sample
MIN_OCCURANCES_ACROSS_PERCENTAGE_OF_SAMPLES = 0.20 # only keep taxa that occurred in at least 20% of samples
```

# Phyloseq Data Object

## Turn CSV into Phyloseq Data Object

```{r}
otu_mat <- as.matrix(read.table(file = paste(raw_location, 'processed_OTU_counts.tsv', sep = ""), sep = '\t', header = TRUE, row.names = 1))
tax_mat <- as.matrix(read.table(file = paste(raw_location, 'processed_taxonomy_table.tsv', sep = ""), sep = '\t', header = TRUE, row.names = 1))
sample_meta <- read.table(file = paste(raw_location, 'processed_sample_metadata.tsv', sep = ""), header = TRUE, sep = '\t', row.names = 1)
sample_meta$day <- factor(sample_meta$day, levels=c("3", "5", "7", "10", "20"))
sample_meta$transfer <- factor(sample_meta$transfe, levels=c("0", "1", "2", "3"))

# get phylogenetic tree
tree <- ape::read.tree(paste(raw_location, "MUSCLE_alignment_ML_tree_all_OTUs", sep=""))

class(otu_mat)
class(tax_mat)
class(sample_meta)
```
```{r}
sample_meta
```

```{r}
OTU = otu_table(otu_mat, taxa_are_rows = TRUE)
TAX = tax_table(tax_mat)
ps = phyloseq(OTU, TAX, tree)
sample_data(ps) <- sample_meta
ps
```
## Cleaning

```{r}
# check that all samples have at least 5k reads
(all(phyloseq::sample_sums(ps) > MIN_SAMPLE_RC))
# only keep taxa that occurred at an abundance of greater than 0.1% in any sample
ps_rel_abund = phyloseq::transform_sample_counts(ps, function(x){x / sum(x)})
(ps <- phyloseq::prune_taxa(rowSums(otu_table(ps_rel_abund) > MIN_PERCENT_IN_ANY_SAMPLE) > 0L
                           , ps))
# only keep taxa that occurred in at least 20% of samples
(ps <- phyloseq::prune_taxa(rowSums(otu_table(ps) > 0L) > length(sample_names(ps)) * MIN_OCCURANCES_ACROSS_PERCENTAGE_OF_SAMPLES
                           , ps))
```

```{r}
# Save
saveRDS(ps, paste(save_location, "phyloseq.rds", sep = ""))
```

```{r}
# Get CLR transform values
(ps_clr <- microbiome::transform(ps, "clr"))
# Save
phyloseq::otu_table(ps_clr)[1:5, 1:5]
write.csv(phyloseq::otu_table(ps_clr), paste(save_location, "OTU_counts_clean_clr.csv", sep = ""))
```

```{r}
# Get a subsection of only samples on cellulose
(ps_cellulose <- phyloseq::subset_samples(ps, carbon == "cellulose" | carbon == "original")) 
(ps_cellulose <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_cellulose) > 0, ps_cellulose))

ps_cellulose %>% 
  sample_data %>%
  dplyr::count(group)

saveRDS(ps_cellulose, paste(save_location, "phyloseq_cellulose.rds", sep=""))
```

---

## Load Phyloseq Data Object

```{r}
ps <- readRDS(paste(save_location, "phyloseq.rds", sep=""))
ps
```
```{r}
sample_data(ps)
```

# Visualize Bar Chart Counts

```{r}
ps %>% 
  sample_data %>%
  dplyr::count(carbon)
```
## Visualize Raw Counts

```{r}
# Plot class
phyloseq::plot_bar(ps, fill = "class") +
  geom_bar(aes(color = class, fill = class), stat = "identity", position = "stack") +
  labs(x = "", y = "Counts\n") +
  facet_wrap(~factor(group, levels=c("C0C", "1C", "2C", "1G", "2G", "3G", "1M", "2M", "3M")), scales = "free") +
  theme(panel.background = element_blank(),
        axis.ticks.x=element_blank())
```

## Visualize Relative Abundances

```{r}
#Get count of class
table(phyloseq::tax_table(ps)[, "class"])
```

```{r}
#Get count of order
table(phyloseq::tax_table(ps)[, "order"])
```
```{r}
#Get count of phylum
table(phyloseq::tax_table(ps)[, "phylum"])
```

```{r}
#Transform to relative abundance
ps_rel_abund = phyloseq::transform_sample_counts(ps, function(x){x / sum(x)})
phyloseq::otu_table(ps)[1:5, 1:5]
phyloseq::otu_table(ps_rel_abund)[1:5, 1:5]
```

```{r}
# Plot genus
phyloseq::plot_bar(ps_rel_abund, fill = "genus") +
  geom_bar(aes(color = class, fill = class), stat = "identity", position = "stack") +
  labs(x = "", y = "Relative Abundance\n") +
  facet_wrap(~factor(group, levels=c("C0C", "1C", "2C", "1G", "2G", "3G", "1M", "2M", "3M")), scales = "free") +
  theme(panel.background = element_blank(),
        axis.ticks.x=element_blank())

# Plot family
phyloseq::plot_bar(ps_rel_abund, fill = "family") +
  geom_bar(aes(color = class, fill = class), stat = "identity", position = "stack") +
  labs(x = "", y = "Relative Abundance\n") +
  facet_wrap(~factor(group, levels=c("C0C", "1C", "2C", "1G", "2G", "3G", "1M", "2M", "3M")), scales = "free") +
  theme(panel.background = element_blank(),
        axis.ticks.x=element_blank())

# Plot order
phyloseq::plot_bar(ps_rel_abund, fill = "order") +
  geom_bar(aes(color = class, fill = class), stat = "identity", position = "stack") +
  labs(x = "", y = "Relative Abundance\n") +
  facet_wrap(~factor(group, levels=c("C0C", "1C", "2C", "1G", "2G", "3G", "1M", "2M", "3M")), scales = "free") +
  theme(panel.background = element_blank(),
        axis.ticks.x=element_blank())

# Plot class
phyloseq::plot_bar(ps_rel_abund, fill = "class") +
  geom_bar(aes(color = class, fill = class), stat = "identity", position = "stack") +
  labs(x = "", y = "Relative Abundance\n") +
  facet_wrap(~factor(group, levels=c("C0C", "1C", "2C", "1G", "2G", "3G", "1M", "2M", "3M")), scales = "free") +
  theme(panel.background = element_blank(),
        axis.ticks.x=element_blank())

# Plot phylum
phyloseq::plot_bar(ps_rel_abund, fill = "phylum") +
  geom_bar(aes(color = phylum, fill = phylum), stat = "identity", position = "stack") +
  labs(x = "", y = "Relative Abundance\n") +
  facet_wrap(~factor(group, levels=c("C0C", "1C", "2C", "1G", "2G", "3G", "1M", "2M", "3M")), scales = "free") +
  theme(panel.background = element_blank(),
        axis.ticks.x=element_blank())
```

## Visualize CLR Transformed

```{r}
# Plot class
phyloseq::plot_bar(ps_clr, fill = "class") +
  geom_bar(aes(color = class, fill = class), stat = "identity", position = "stack") +
  labs(x = "", y = "Centered Log Ratio\n") +
  facet_wrap(~factor(group, levels=c("C0C", "1C", "2C", "1G", "2G", "3G", "1M", "2M", "3M")), scales = "free") +
  theme(panel.background = element_blank(),
        axis.ticks.x=element_blank())
```
# Correlation Clustering

## Aitchison Distance PCA

```{r}
plot_clr_pca <- function(ps, taxrank="none", color="cluster") {
  #PCA via phyloseq
  if (taxrank == "none") {
    ord_clr <- ps %>%
      microbiome::transform("clr") %>%
      phyloseq::ordinate("RDA")
  } else {
    ord_clr <- ps %>%
      tax_glom(taxrank = taxrank, NArm = TRUE) %>% 
      microbiome::transform("clr") %>%
      phyloseq::ordinate("RDA")
  }
  
  #Scale axes and plot ordination
  clr1 <- ord_clr$CA$eig[1] / sum(ord_clr$CA$eig)
  clr2 <- ord_clr$CA$eig[2] / sum(ord_clr$CA$eig)
  phyloseq::plot_ordination(ps, ord_clr, type="samples", color=color) +
    geom_point(size = 2) +
    coord_fixed(clr2 / clr1) +
    stat_ellipse(aes(group = carbon), linetype = 2)
  
  # #Plot scree plot
  # phyloseq::plot_scree(ord_clr) + 
  #   geom_bar(stat="identity", fill = "blue") +
  #   labs(x = "\nAxis", y = "Proportion of Variance\n")
}
```

```{r}
plot_clr_pca(ps, "none", color="group")
plot_clr_pca(ps, "none", color="transfer")
plot_clr_pca(ps, "none", color="day")
```

```{r}
plot_clr_pca(ps, "species", color="group")
plot_clr_pca(ps, "species", color="transfer")
plot_clr_pca(ps, "species", color="day")
```

```{r}
plot_clr_pca(ps, "genus", color="group")
plot_clr_pca(ps, "genus", color="transfer")
plot_clr_pca(ps, "genus", color="day")
```

```{r}
plot_clr_pca(ps, "family", color="group")
plot_clr_pca(ps, "family", color="transfer")
plot_clr_pca(ps, "family", color="day")
```

## Pretty PCA Plots

```{r}
plot_pretty_clr_pca <- function(ps, taxrank) {
  #PCA via phyloseq
  if (taxrank == "none") {
    ord_clr <- ps %>%
      microbiome::transform("clr") %>%
      phyloseq::ordinate("RDA")
  } else {
    ord_clr <- ps %>%
      tax_glom(taxrank = taxrank, NArm = TRUE) %>% 
      microbiome::transform("clr") %>%
      phyloseq::ordinate("RDA")
  }
  
  # make data frame
  clr_pca.data <- data.frame(PC1=ord_clr$CA$u[,1], PC2=ord_clr$CA$u[,2])
  meta <- data.frame(phyloseq::sample_data(ps))
  clr_pca.data <- merge(clr_pca.data, meta, by="row.names")
  clr_pca.data$day <- as.factor(clr_pca.data$day)
  
  #Scale axes and plot ordination
  clr1 <- ord_clr$CA$eig[1] / sum(ord_clr$CA$eig)
  clr2 <- ord_clr$CA$eig[2] / sum(ord_clr$CA$eig)
  ggplot(clr_pca.data, aes(x = PC1, y = PC2)) + 
    theme_classic() + 
    geom_point(
      mapping = aes(colour = day, shape = carbon),
      size = 3,
      alpha = 5/6
    ) +
    coord_fixed(clr2 / clr1) +
    stat_ellipse(aes(group = carbon), linetype = 2) +
    scale_color_manual(values = c(
    "3" = "#B2E5EA", # lightest
    "5" = "#00BAE7",
    "7" = "#0E6FA0",
    "10" = "#012652",
    "20" = "#01091C" # darkest
  )) +
    scale_shape_manual(values = c(
    "original" = 4,
    "cellulose" = 16,
    "glucose" = 18,
    "malate" = 15
  )) +
    xlab(sprintf("PC 1 %s%%", round(100 * clr1, 2))) +
    ylab(sprintf("PC 2 %s%%", round(100 * clr2, 2))) +
    ggtitle(sprintf("PCA taxrank: %s", taxrank))
  
  # #Plot scree plot
  # phyloseq::plot_scree(ord_clr) + 
  #   geom_bar(stat="identity", fill = "blue") +
  #   labs(x = "\nAxis", y = "Proportion of Variance\n")
}


plot_pretty_clr_pca_detailed <- function(ps, taxrank) {
  #PCA via phyloseq
  if (taxrank == "none") {
    ord_clr <- ps %>%
      microbiome::transform("clr") %>%
      phyloseq::ordinate("RDA")
  } else {
    ord_clr <- ps %>%
      tax_glom(taxrank = taxrank, NArm = TRUE) %>% 
      microbiome::transform("clr") %>%
      phyloseq::ordinate("RDA")
  }
  
  # make data frame
  clr_pca.data <- data.frame(PC1=ord_clr$CA$u[,1], PC2=ord_clr$CA$u[,2])
  meta <- data.frame(phyloseq::sample_data(ps))
  clr_pca.data <- merge(clr_pca.data, meta, by="row.names")
  clr_pca.data$day <- as.factor(clr_pca.data$day)
  
  #Scale axes and plot ordination
  clr1 <- ord_clr$CA$eig[1] / sum(ord_clr$CA$eig)
  clr2 <- ord_clr$CA$eig[2] / sum(ord_clr$CA$eig)
  ggplot(clr_pca.data, aes(x = PC1, y = PC2)) + 
    theme_classic() + 
    geom_point(
      mapping = aes(colour = day, shape = group),
      size = 3,
      alpha = 5/6
    ) +
    coord_fixed(clr2 / clr1) +
    stat_ellipse(aes(group = carbon), linetype = 2) +
    scale_color_manual(values = c(
    "3" = "#B2E5EA", # lightest
    "5" = "#00BAE7",
    "7" = "#0E6FA0",
    "10" = "#012652",
    "20" = "#01091C" # darkest
  )) +
    scale_shape_manual(values = c(
    "C0C" = 4,
    "1C" = 10,
    "2C" = 16,
    "2G" = 9,
    "3G" = 18,
    "2M" = 12,
    "3M" = 15
  )) +
    xlab(sprintf("PC 1 %s%%", round(100 * clr1, 2))) +
    ylab(sprintf("PC 2 %s%%", round(100 * clr2, 2))) +
    ggtitle(sprintf("PCA taxrank: %s", taxrank))
  
  # #Plot scree plot
  # phyloseq::plot_scree(ord_clr) + 
  #   geom_bar(stat="identity", fill = "blue") +
  #   labs(x = "\nAxis", y = "Proportion of Variance\n")
}
```


```{r}
plot_pretty_clr_pca(ps, taxrank = "none")
plot_pretty_clr_pca(ps, taxrank = "species")
plot_pretty_clr_pca(ps, taxrank = "genus")
```

Check that results are robust to sub-sampling only samples on cellulose

```{r}
plot_pretty_clr_pca_detailed(ps, taxrank = "none")
plot_pretty_clr_pca_detailed(ps, taxrank = "species")
plot_pretty_clr_pca_detailed(ps, taxrank = "genus")
plot_pretty_clr_pca_detailed(ps, taxrank = "family")
```

```{r}
plot_pretty_clr_pca_detailed(ps_cellulose, taxrank = "none")
plot_pretty_clr_pca_detailed(ps_cellulose, taxrank = "species")
plot_pretty_clr_pca_detailed(ps_cellulose, taxrank = "genus")
```

## CLR --> Ward Distance --> Clustering Dendogram

```{r}
get_ward <- function(ps, taxrank) {
  #PCA via phyloseq
  if (taxrank == "none") {
    ps_clr <- ps %>%
      microbiome::transform("clr")
  } else {
    ps_clr <- ps %>%
      tax_glom(taxrank = taxrank, NArm = TRUE) %>% 
      microbiome::transform("clr")
  }
  
  #Get Aitchison distance matrix
  ps_clr_otu <- data.frame(phyloseq::otu_table(ps_clr))
  ps_clr_otu <- t(ps_clr_otu)
  a_dist_clr <- vegan::vegdist(ps_clr_otu,  method = "euclidean")
  
  #Save as dendrogram
  ward <- as.dendrogram(hclust(a_dist_clr, method = "ward.D2"))
  
  #Provide color codes
  meta <- data.frame(phyloseq::sample_data(ps))
  colorCode <- c(C = "green", M = "orange", G = "blue")
  labels_colors(ward) <- colorCode[meta$series][order.dendrogram(ward)]
  
  #Plot color by code
  par(cex = 0.4)
  ward %>% plot(main = sprintf("Aitchison on Centered Log-Ratios \ntaxrank: %s", taxrank))
  
  #Plot color by cluster
  ward %>% set("labels_cex", 1) %>% set("labels_col", value = c(1, 2, 3), k=3) %>% 
     plot(main = sprintf("Aitchison: Color labels per cluster \ntaxrank: %s", taxrank))
  
  #Plot color by cluster
  ward %>% set("labels_cex", 1) %>% set("labels_col", value = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), k=9) %>% 
     plot(main = sprintf("Aitchison: Color labels per cluster \ntaxrank: %s", taxrank))
  
  return(ward)
}
```

```{r}
ward_cellulose_OTU <- get_ward(ps_cellulose, "none")
```

```{r}
ward_cellulose_species <- get_ward(ps_cellulose, "species")
```

```{r}
ward_cellulose_genus <- get_ward(ps_cellulose, "genus")
```
```{r}
ward_OTU <- get_ward(ps, "none")
```
```{r}
ward_species <- get_ward(ps, "species")
```



```{r}
(clusters <- ward_cellulose_OTU %>% set("labels_col", value = c("late", "late", "mid"), k=3) %>% labels_colors)
write.csv(clusters, file=paste(save_location, "cluster_decisions.csv", sep=""), quote=FALSE)
```

# Appendix

## Relative Abundance --> Bray-Curtis Distance --> PCoA & Dendogram

```{r}
#PCA via phyloseq
ord_bc <- phyloseq::ordinate(ps_rel_abund, "PCoA", "bray")
bc_pca.data <- data.frame(PC1=ord_bc$vectors[,1], PC2=ord_bc$vectors[,2])
meta <- data.frame(phyloseq::sample_data(ps_rel_abund))
(bc_pca.data <- merge(bc_pca.data, meta, by="row.names"))
bc_pca.data$day <- as.factor(bc_pca.data$day)
#Plot scree plot
phyloseq::plot_scree(ord_bc) + 
  geom_bar(stat="identity", fill = "blue") +
  labs(x = "\nAxis", y = "Proportion of Variance\n")

# Examine eigenvalues and % prop. variance explained
head(ord_bc$values$Eigenvalues)   
sapply(ord_bc$values$Eigenvalues[1:5], function(x) x / sum(ord_bc$values$Eigenvalues))

# plot
clr1 <- ord_bc$values$Eigenvalues[1] / sum(ord_bc$values$Eigenvalues)
clr2 <- ord_bc$values$Eigenvalues[2] / sum(ord_bc$values$Eigenvalues)
ggplot(bc_pca.data, aes(x = PC1, y = PC2)) + 
  theme_classic() + 
  geom_point(
    mapping = aes(colour = day, shape = carbon),
    size = 2,
    alpha = 5/6
  ) +
  coord_fixed(clr2 / clr1) +
  stat_ellipse(aes(group = carbon), linetype = 2) +
  scale_color_manual(values = c(
  "3" = "#03045e",
  "5" = "#0077b6",
  "7" = "#00b4d8",
  "10" = "#90e0ef",
  "20" = "#caf0f8"
)) +
  scale_shape_manual(values = c(
  "original" = 4,
  "cellulose" = 16,
  "glucose" = 18,
  "malate" = 15
)) +
  xlab(sprintf("PC 1 [%s%%]", round(100 * clr1, 2))) +
  ylab(sprintf("PC 2 [%s%%]", round(100 * clr2, 2)))
```


```{r}
# Generate distance matrix
ps_cellulose_rel_abund <- phyloseq::subset_samples(ps, carbon =="cellulose")
ps_cellulose_rel_abund <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_cellulose_rel_abund) > 0, ps_cellulose_rel_abund)
ord_bc_cellulose <- phyloseq::ordinate(ps_cellulose_rel_abund, "PCoA", "bray")
bc_pca_cellulose.data <- data.frame(PC1=ord_bc_cellulose$vectors[,1], PC2=ord_bc_cellulose$vectors[,2])
meta <- data.frame(phyloseq::sample_data(ps_cellulose_rel_abund))
(bc_pca_cellulose.data <- merge(bc_pca_cellulose.data, meta, by="row.names"))
bc_pca_cellulose.data$day <- as.factor(bc_pca_cellulose.data$day)
#Plot scree plot
phyloseq::plot_scree(ord_bc_cellulose) + 
  geom_bar(stat="identity", fill = "blue") +
  labs(x = "\nAxis", y = "Proportion of Variance\n")

# plot
clr1 <- ord_bc_cellulose$values$Eigenvalues[1] / sum(ord_bc_cellulose$values$Eigenvalues)
clr2 <- ord_bc_cellulose$values$Eigenvalues[2] / sum(ord_bc_cellulose$values$Eigenvalues)
ggplot(bc_pca_cellulose.data, aes(x = PC1, y = PC2)) + 
  theme_classic() + 
  geom_point(
    mapping = aes(colour = day, shape = group),
    size = 3,
    alpha = 5/6
  ) +
  coord_fixed(clr2 / clr1) +
  stat_ellipse(aes(group = carbon), linetype = 2) +
  scale_color_manual(values = c(
  "3" = "#03045e",
  "5" = "#0077b6",
  "7" = "#00b4d8",
  "10" = "#90e0ef",
  "20" = "#caf0f8"
)) +
  scale_shape_manual(values = c(
  "C0C" = 4,
  "1C" = 16,
  "2C" = 10,
  "2G" = 18,
  "3G" = 9,
  "2M" = 15,
  "3M" = 12
)) +
  xlab(sprintf("PC 1 [%s%%]", round(100 * clr1, 2))) +
  ylab(sprintf("PC 2 [%s%%]", round(100 * clr2, 2)))
```

```{r}
#Extract OTU table and compute BC
ps_rel_otu <- data.frame(phyloseq::otu_table(ps_rel_abund))
ps_rel_otu <- t(ps_rel_otu)
bc_dist <- vegan::vegdist(ps_rel_otu, method = "bray")
as.matrix(bc_dist)[1:5, 1:5]
```

```{r}
#Save as dendrogram
ward <- as.dendrogram(hclust(bc_dist, method = "ward.D2"))
#Provide color codes
meta <- data.frame(phyloseq::sample_data(ps_rel_abund))
colorCode <- c(sC = "green", sM = "orange", sG = "blue")
labels_colors(ward) <- colorCode[meta$series][order.dendrogram(ward)]
#Plot
par(cex = 0.4)
ward %>% plot(main = "Bray-Curtis on Relative Abundance")

# color by cluster
ward %>% set("labels_cex", 1) %>% set("labels_col", value = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), k=9) %>% 
   plot(main = "Bray-Curtis: Color labels \nper cluster")
```

## Using UniFrac

```{r}
#Generate distances
ord_unifrac <- ordinate(ps_cellulose, method = "PCoA", distance = "wunifrac") 
ord_unifrac_un <- ordinate(ps_cellulose, method = "PCoA", distance = "unifrac")   
#Plot ordinations
a <- plot_ordination(ps, ord_unifrac, color = "day") + geom_point(size = 2)
b <- plot_ordination(ps, ord_unifrac_un, color = "day") + geom_point(size = 2)
cowplot::plot_grid(a, b, nrow = 1, ncol = 2, scale = .9, labels = c("Weighted", "Unweighted"))
```
