---
title: "visualize differential abundance"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

Goal: since there is a significant difference in the composition of clusters, find out who is driving the difference

Groups
* "late" vs. "mid"

* "perturbed M" vs. "perturbed G"

* "mid C" vs. "mid G"
* "mid C" vs. "mid M"
* "mid G" vs. "mid M"

* "late C" vs. "late G" 
* "late C" vs. "late M"
* "late G" vs. "late M"

* "late C" vs. all the others

Methods:
* Wilcox Sum Rank Test
* ALDEx2
* ANCOM-BC

Levels
* Phylum
* Class
* Order
* Family
* Genus
* Species
* OTU-level

# Install Packages

```{r}
# .cran_packages <- c("tidyverse", "cowplot", "picante", "vegan", "HMP", "dendextend", "rms", "devtools")
# .inst <- .cran_packages %in% installed.packages()
# if(any(!.inst)) {
#   install.packages(.cran_packages[!.inst])
# }
```

```{r}
# .bioc_packages <- c("phyloseq", "Biostrings", "DESeq2", "microbiome", "metagenomeSeq", "ALDEx2")
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(.bioc_packages, version = "3.14")
```


# Loading Required Packages

```{r}
library("ggplot2"); packageVersion("ggplot2") # plotting
```
```{r}
library("tidyverse"); packageVersion("tidyverse") # util              
```

```{r}
library("dplyr"); packageVersion("dplyr") # util
```

```{r}
library("phyloseq"); packageVersion("phyloseq") # data object                 
```

```{r}
library("ape"); packageVersion("ape") # taxa string manipulation
```

```{r}
library(microbiome); packageVersion("microbiome") # clr transformation
```

```{r}
library(dendextend); packageVersion("dendextend") # extend customization of dendograms
```

```{r}
library(vegan); packageVersion("vegan") # bray-curtis and other metrics
```

```{r}
# Manipulation and visualization of taxonomic data, particularly those from amplicon metagenomics research. (Foster, Sharpton, and Grünwald 2017)
library(metacoder); packageVersion("metacoder")
```

```{r}
# defines taxonomic classes and functions to manipulate them. The goal is to use these classes as low level fundamental taxonomic classes that other R packages can build on and use. This is used by metacoder. 
library(taxa); packageVersion("taxa")
```

# Initialization

```{r}
# ggplot2 package theme set
theme_set(theme_void())

# control randomization
set.seed(42)

# save outputs
save_location = "../data/processed_FCF/"
```

# Load Phyloseq Data Object

```{r}
(ps <- readRDS(paste(save_location, "phyloseq.rds", sep="")))
# convert to taxmap format 
taxmap <- parse_phyloseq(ps)
taxmap
```

# Plot metacoder family tree

```{r}
set.seed(42)
heat_tree(taxmap,
          node_label = taxon_names,
          node_size = n_obs,
          node_color = n_obs,
          node_color_axis_label = "OTU count",
          layout = "da",
          initial_layout = "re",
          # output_file = "../data/processed_FCF_53OTUs/metacoder_differential_abundance/FCF.png"
          title = "FCF")
```




# Extract tax_ids

```{r}
taxmap.key <- merge(
  as.data.frame(taxon_names(taxmap)), 
  as.data.frame((taxon_ranks(taxmap))), 
  by.x=0, by.y=0)
colnames(taxmap.key) <- c("taxon_id", "taxon_names", "taxon_rank")
taxmap.key
```

```{r}
#Print number of duplicated rows 
#i.e. taxon not uniquely identified by taxon_name + taxon_rank
nrow(taxmap.key) - nrow(
  taxmap.key[!(duplicated(taxmap.key[c("taxon_names","taxon_rank")]) 
             | duplicated(taxmap.key[c("taxon_names","taxon_rank")],
            fromLast = TRUE)), ])
```

# Process DA .csv

```{r}
get_taxon_id = function(x) {
  #Get taxrank (finest resolution taxonomy)
  taxrank <- x[["taxrank"]]
  #Get tax name from relevant column
  taxname <- x[[taxrank]]
  
  #Query on tax_data
  if (taxrank == "OTU") {
    matches <- filter(taxmap$data$tax_data, otu_id == taxname)
    return(matches[1]$taxon_id)
  }
  
  #Query on taxmap.key
  matches <- filter(taxmap.key, taxon_names == taxname & taxon_rank == taxrank)
  return(matches[1]$taxon_id)
}
```

## ALDEx2

```{r}
#Read .csv
ALDEx2_results <- read.table(
  file = paste(save_location, 'ALDEx2_results.csv', sep=""), 
  header = TRUE, sep = ',')
#Get taxon_id
ALDEx2_results <- cbind(ALDEx2_results, taxon_id = apply(
  ALDEx2_results, 1, get_taxon_id))
#Zero effect size for non-significant p-values
ALDEx2_results$aldex2_effect[ALDEx2_results$aldex2_wi.eBH > 0.05] <- 0
#Show final table
ALDEx2_results
```

## Get Differential Abundance Table

```{r}
get_da_taxmap = function(taxmap, ti) {
  ALDEx2_results_filtered <- filter(ALDEx2_results, 
                                    test_id == ti & taxrank != "OTU")
  
  #Merge Dataframe
  taxmap$data$da <- taxmap.key %>%
    merge(ALDEx2_results_filtered, by.x = "taxon_id", by.y = "taxon_id", all.x = TRUE) %>%
    dplyr::select("taxon_id", "aldex2_diff.btw", "aldex2_diff.win", "aldex2_effect", "aldex2_positive", "aldex2_negative") %>%
    mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>%
    as.tibble()
  
  return(taxmap)
}
```

# Heat Map Trees

```{r}
plot_comparative_heat_tree = function(ti) {
  
  #Define color key
  focused.cluster_levels <- c("late C", "late G", "late M",
                            "mid C", "mid G", "mid M",
                            "perturbed G", "perturbed M")

  focused.cluster_colors <- c("springgreen4", "steelblue4", "sienna4",
                              "springgreen3", "steelblue3", "sienna3",
                              "steelblue1", "sienna1")
  names(focused.cluster_colors) <- focused.cluster_levels

  #Get taxmap
  tmp <- get_da_taxmap(taxmap, ti)
  group1 <- unique(tmp$data$da$aldex2_positive)
  group1 <- group1[!is.na(group1)][1];
  group2 <- unique(tmp$data$da$aldex2_negative)
  group2 <- group2[!is.na(group2)][1];
  group1
  group2

  #Get corresponding colors
  blank_color <- "oldlace"
  # blank_color <- "gray"
  g1_color <- focused.cluster_colors[[group1]]
  g2_color <- focused.cluster_colors[[group2]]
  
  #Get min, max
  aldex2_min <- min(tmp$data$da$aldex2_effect)
  aldex2_max <- max(tmp$data$da$aldex2_effect)
  aldex2_range <- max(abs(aldex2_min), abs(aldex2_max))
  
  # ALDEx2
  # control randomization
  set.seed(42)
  tmp %>%
    filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
    heat_tree(node_size = n_obs, # number of OTUs
              node_size_axis_label = "OTU count",
              node_color = aldex2_effect, # difference between groups
              node_color_interval = c(-1 * aldex2_range, aldex2_range), # symmetric interval
              node_color_range = c(g2_color, blank_color, g1_color), # diverging colors
              node_color_axis_label = "ALDEx2 Effect Size",
              # node_label = taxon_names, # comment to remove labels
              initial_layout = "reingold-tilford",
              layout = "davidson-harel",
              title = paste(ti, "positive to", group1, "ALDEx2"),
              output_file = paste(save_location,
                paste("metacoder_differential_abundance/heat_tree", 
                ti, "ALDEX2.png", sep="_"), sep=""))
  
}
```

```{r}
unique(ALDEx2_results$test_id)
```

```{r}
unique(ALDEx2_results$aldex2_positive)
```
```{r}
for (test_id in unique(ALDEx2_results$test_id)) {
  tryCatch(
    {
      plot_comparative_heat_tree(test_id)
      print(test_id)
    },
    error=function(cond) {
      message(test_id)
      message("-failed to run")
    }
  )  
}
```
## Get one with labels
```{r}
# empty
# plot_comparative_heat_tree("mG_vs_mM")
```

# Scatter plots

```{r}
get_da_scatter_ALDEx2 <- function(data, test_ids = c(), taxrank = "genus", plot_name="x") {
  # only use OTU level differential analysis
  # filter and color data points
  ALDEx2_results_filtered <- data %>%
    filter(test_id %in% test_ids) %>% 
    filter(taxrank == "OTU") %>% 
    # only plot significant OTUs
    filter(aldex2_effect != 0) %>%
    mutate(significant = case_when(
      aldex2_wi.eBH < 0.05 & aldex2_effect > 1 ~ aldex2_positive,
      aldex2_wi.eBH < 0.05 & aldex2_effect < -1 ~ aldex2_negative,
      TRUE~"not significant"))
  
  
  # get number of y-axis categories
  num_categories <- length(unique(ALDEx2_results_filtered[[taxrank]]))
  
  # get range of x-axis
  min_effect <- max(ALDEx2_results_filtered$aldex2_effect)
  max_effect <- min(ALDEx2_results_filtered$aldex2_effect)
  range_effect <- max(abs(min_effect), abs(max_effect))
  
  # divide taxrank categories into strips (for zebra striped background)
  strip_data <- data.frame(unique(ALDEx2_results_filtered[[taxrank]]))
  colnames(strip_data) <- c("taxrank")
  strip_data <- strip_data %>%
    arrange(desc(taxrank)) %>%
    mutate(xmin = -1 * range_effect, xmax = range_effect,
           y_position = rev(1:nrow(.)),
           ymin = y_position - 0.5,
           ymax = y_position + 0.5,
           fill = rep(c("red", "blue"), length.out=nrow(strip_data)))
  
  # map taxrank to strip position
  refer_y_pos <- as.numeric(strip_data$y_position)
  names(refer_y_pos) <- strip_data$taxrank
  ALDEx2_results_filtered$y_position <- as.numeric(refer_y_pos[ALDEx2_results_filtered[[taxrank]]])
  ALDEx2_results_filtered
  
  # pivot strip data table
  strip_data <- strip_data %>%
    pivot_longer(cols=c(xmin, xmax), values_to="x", names_to="xmin_xmax") %>%
    dplyr::select(-xmin_xmax)

  # plot pretty
  g <- ALDEx2_results_filtered %>%
    ggplot(axes = FALSE, show.legend = FALSE,
           aes_string(x = "aldex2_effect", y = "y_position", colour = "significant", group="y_position", alpha=0.01)) +
    # zebra stripe background for readability
    geom_ribbon(data=strip_data,
                inherit.aes = FALSE, show.legend = FALSE,
                aes(x=x, ymin=ymin, ymax=ymax, group=y_position, fill=fill)) +
    scale_fill_manual(name = 'fill', values = c("gray100", "gray95")) +
    scale_colour_manual(name = 'significant', 
                        values = setNames(c("gray50", 
                                            "black", "blue", "green",
                                            "black", "slateblue1", "sienna1",
                                            "springgreen3", "slateblue3", "sienna3", 
                                            "springgreen4", "slateblue4", "sienna4"),
                                          c("not significant", 
                                            "initial", "mid", "late",
                                            "early", "perturbed G", "perturbed M",
                                            "mid C", "mid G", "mid M",
                                            "late C", "late G", "late M")
                            )) + 
    geom_jitter(width = 0, height = 0.10) + 
    facet_grid(cols = vars(test_id)) + 
    scale_y_discrete(taxrank, limits=rev(names(refer_y_pos))) +
    scale_x_discrete("Effect Size", limits=c(-1, 0, 1)) + 
    geom_vline(xintercept=1, linetype="dotted", color="gray70", size=1) +
    geom_vline(xintercept=-1, linetype="dotted", color="gray70", size=1) +
    theme_set(theme_classic()) + 
    theme(legend.position = "none",
          panel.spacing = unit(0.5, "lines"),
          panel.border = element_rect(color = "black", fill = NA, size = 1.5), 
          strip.background = element_rect(color = "black", size = 1))
  print(g)  
}
```

```{r}
unique(ALDEx2_results$test_id)
```

## Cluster level

```{r}
test_ids <- c("p_vs_m", "p_vs_l", "p_vs_m")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "genus")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "order")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "class")
```
## Focused Cluster level

```{r}
test_ids <- c("pG_vs_pM", "mC_vs_mM", "mC_vs_mG", "mG_vs_mM", "lC_vs_lM", "lC_vs_lG", "lG_vs_lM")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "genus", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "order", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "class", plot_name=plot_name)
```

```{r}
test_ids <- c("lC_vs_lM", "lC_vs_lG", "lC_vs_pM", "lC_vs_pG", "lC_vs_mC", "lC_vs_mG", "lC_vs_mM")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "genus", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "order", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "class", plot_name=plot_name)
```








