---
title: "differential abundance aldex2"
output: html_notebook
---

Goal: since there is a significant difference in the composition of clusters, find out who is driving the difference

Methods:
* Wilcox Sum Rank Test
* ALDEx2
* ANCOM-BC

Levels
* Phylum
* Class
* Order
* Family
* Genus
* Species
* OTU

# Install Packages

```{r}
# .cran_packages <- c("tidyverse", "cowplot", "picante", "vegan", "HMP", "dendextend", "rms", "devtools")
# .inst <- .cran_packages %in% installed.packages()
# if(any(!.inst)) {
#   install.packages(.cran_packages[!.inst])
# }
```

```{r}
# .bioc_packages <- c("phyloseq", "Biostrings", "DESeq2", "microbiome", "metagenomeSeq", "ALDEx2", "ANCOMBC")
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(.bioc_packages, version = "3.14")
```

# Loading Required Packages

```{r}
library("ggplot2"); packageVersion("ggplot2") # plotting
```
```{r}
library("tidyverse"); packageVersion("tidyverse") # util              
```
```{r}
library("dplyr"); packageVersion("dplyr") # util
```
```{r}
library("phyloseq"); packageVersion("phyloseq") # data object                 
```
```{r}
library("ape"); packageVersion("ape") # taxa string manipulation
```
```{r}
library(microbiome); packageVersion("microbiome") # clr transformation
```
```{r}
library(dendextend); packageVersion("dendextend") # extend customization of dendograms
```
```{r}
library(vegan); packageVersion("vegan") # bray-curtis and other metrics
```

```{r}
library(ALDEx2); packageVersion("ALDEx2") # aldex2 test (compositional differential abundance)
```
# Initialization

```{r}
# ggplot2 package theme set
theme_set(theme_bw())

# control randomization
set.seed(42)

# save outputs
save_location = "../data/processed_carbon_53OTUs/"
```

# Aes

```{r}
# color - days
color_day = c(
    "3" = "#B2E5EA", # lightest
    "5" = "#00BAE7",
    "7" = "#0E6FA0",
    "10" = "#012652",
    "20" = "#01091C"
    )

# color - growth.phase
color_growth.phase = c(
  "early" = "black",
  "mid C:C" = "springgreen3",
  "late C:C" = "springgreen4",
  "perturbed G" = "steelblue1",
  "mid G:C" = "steelblue3",
  "late G:C" = "steelblue4",
  "perturbed M" = "sienna1",
  "mid M:C" = "sienna3",
  "late M:C" = "sienna4"
  )

# shape - series
shape_series = c(
  "C0C" = 4,
  "C" = 16,
  "G" = 15,
  "M" = 17
  )
# shape - carbon
shape_carbon = c(
  "original" = 4,
  "cellulose" = 16,
  "glucose" = 18,
  "malate" = 15
)
# shape - transfer
shape_transfer = c(
    "C0C" = 4,
    "1C" = 10,
    "2C" = 16,
    "1G" = 0,
    "2G" = 12,
    "3G" = 15,
    "1M" = 5,
    "2M" = 9,
    "3M" = 18
  )
# shape - growth phase
shape_growth.phase = c(
  "early" = 1,
  "mid C:C" = 16,
  "late C:C" = 16,
  "perturbed G" = 15,
  "mid G:C" = 15,
  "late G:C" = 15,
  "perturbed M" = 17,
  "mid M:C" = 17,
  "late M:C" = 17
  )

# color - genus
```

# Load Phyloseq Data Object
```{r}
(ps <- readRDS(paste(save_location, "phyloseq.rds", sep="")))

# Get Relative Abundance transform values
ps_rel_abund = phyloseq::transform_sample_counts(ps, function(x){x / sum(x)})

# Get CLR transform values
ps_clr <- microbiome::transform(ps, "clr")
```

# ALDEx2 Test

```{r}
# data frame to store test results
ALDEx2_results <- data.frame(matrix(nrow = 0, ncol = 4))

# ALDEx2_results <- read.csv2(paste(save_location, "ALDEx2_results.csv", sep=""),
#                            header=TRUE,
#                            sep=",")
# ALDEx2_results$X <- NULL
# ALDEx2_results
```


## Example Output
```{r}
psa <- phyloseq::subset_samples(ps, cluster == "perturbed" | cluster == "mid")

aldex2_da <- ALDEx2::aldex(
    data.frame(phyloseq::otu_table(psa)), 
    phyloseq::sample_data(psa)$cluster, 
    mc.samples = 1000,
    test="t", 
    effect = TRUE, 
    denom="iqlr")
```

## Helper Functions

```{r}
aldex2_on_cluster <- function(ps, test_id) {
  #Run ALDEx2
  aldex2_da <- ALDEx2::aldex(
    data.frame(phyloseq::otu_table(ps)), 
    phyloseq::sample_data(ps)$cluster, 
    mc.samples = 1000,
    test="t", 
    effect = TRUE, 
    denom="iqlr")
  
  #Note positive and negative treatment variable 
  positive <- substr(
    colnames(aldex2_da)[2], 
    start = 9, 
    stop = nchar(colnames(aldex2_da)[2])
  )
  
  negative <- substr(
    colnames(aldex2_da)[3], 
    start = 9, 
    stop = nchar(colnames(aldex2_da)[3])
  )
  
  #Plot effect sizes
  png(file=paste(save_location, "aldex2_differential_abundance/", 
                 test_id, ".png", sep=""))
  par(mfrow=c(1,2))
  ALDEx2::aldex.plot(aldex2_da, type="MA", test="wilcox")
  ALDEx2::aldex.plot(aldex2_da, type="MW", test="wilcox")
  dev.off()
  
  #Clean up presentation
  colnames(aldex2_da) <- paste("aldex2", colnames(aldex2_da), sep = "_")
  sig_aldex2 <- aldex2_da %>%
    rownames_to_column(var = "OTU") %>%
    dplyr::select(OTU, aldex2_diff.btw, aldex2_diff.win, 
                  aldex2_effect, aldex2_wi.ep, aldex2_wi.eBH)
  
  #Get significant names and values to data frame
  sig_aldex2 <- merge(
    x=sig_aldex2, 
    y=phyloseq::tax_table(ps), 
    by.x="OTU", by.y=0, all.x=TRUE)
  
  #Save treatment variable
  sig_aldex2$aldex2_positive <- positive
  sig_aldex2$aldex2_negative <- negative
  
  return(sig_aldex2)
}

aldex2_on_growth_phase <- function(ps, test_id) {
  #Run ALDEx2
  aldex2_da <- ALDEx2::aldex(
    data.frame(phyloseq::otu_table(ps)), 
    phyloseq::sample_data(ps)$growth.phase,
    mc.samples = 1000,
    test="t", 
    effect = TRUE, 
    denom="iqlr")
  
  #Note positive and negative treatment variable 
  positive <- substr(
    colnames(aldex2_da)[2], 
    start = 9, 
    stop = nchar(colnames(aldex2_da)[2])
  )
  
  negative <- substr(
    colnames(aldex2_da)[3], 
    start = 9, 
    stop = nchar(colnames(aldex2_da)[3])
  )
  
  #Plot effect sizes
  png(file=paste(save_location, "aldex2_differential_abundance/", 
                 test_id, ".png", sep=""))
  par(mfrow=c(1,2))
  ALDEx2::aldex.plot(aldex2_da, type="MA", test="wilcox")
  ALDEx2::aldex.plot(aldex2_da, type="MW", test="wilcox")
  dev.off()
  
  #Clean up presentation
  colnames(aldex2_da) <- paste("aldex2", colnames(aldex2_da), sep = "_")
  sig_aldex2 <- aldex2_da %>%
    rownames_to_column(var = "OTU") %>%
    dplyr::select(OTU, aldex2_diff.btw, aldex2_diff.win, 
                  aldex2_effect, aldex2_wi.ep, aldex2_wi.eBH)
  
  #Get significant names and values to data frame
  sig_aldex2 <- merge(
    x=sig_aldex2, 
    y=phyloseq::tax_table(ps), 
    by.x="OTU", by.y=0, all.x=TRUE)
  
  #Save treatment variable
  sig_aldex2$aldex2_positive <- positive
  sig_aldex2$aldex2_negative <- negative
  
  return(sig_aldex2)
}
```

## Drivers
* test at each taxrank

```{r}
rank_names(ps)
```

```{r}
drive_aldex2_on_cluster <- function(ps, test_id) {
  # data frame to store test results
  results <- data.frame(matrix(nrow = 0, ncol = 4))
  
  # prune empty taxa
  ps <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps) > 0, ps)
  
  tryCatch(
      {
        # perform once at OTU level
        rows <- aldex2_on_cluster(ps, paste(test_id, "OTU", sep="_"))
        if (nrow(rows) > 0) {
          rows$taxrank <- "OTU"
          results <- rbind(results, rows) 
        }
        print("OTU") # logging
      },
      error = function(cond) {
        message("OTU-level TEST FAILED!")
        message(cond)
      }
    )
  
  # aggregate at each taxrank
  for (taxrank in rev(rank_names(ps)[-1])) {
    ps.taxa <- tax_glom(ps, taxrank = taxrank, NArm = TRUE)
    # skip if less than 10 taxa
    if (nrow(tax_table(ps.taxa)) < 11) {
      next
    }
    tryCatch(
      {
        # get result of test
        rows <- aldex2_on_cluster(ps.taxa, paste(test_id, taxrank, sep="_"))
        if (nrow(rows) > 0) {
          rows$taxrank <- taxrank
          results <- rbind(results, rows) 
        }
        print(taxrank) # logging
      },
      error = function(cond) {
        message("TEST FAILED!")
        message(cond)
      }
    )
  }
  
  if (nrow(results) > 0) {
    results$test_id <- test_id
    results$level <- "cluster"
  }
  
  return(results)
}
```


```{r}
drive_aldex2_on_growth_phase <- function(ps, test_id) {
  # data frame to store test results
  results <- data.frame(matrix(nrow = 0, ncol = 4))
  
  # prune empty taxa
  ps <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps) > 0, ps)
  
  tryCatch(
      {
        # perform once at OTU level
        rows <- aldex2_on_growth_phase(ps, paste(test_id, "OTU", sep="_"))
        if (nrow(rows) > 0) {
          rows$taxrank <- "OTU"
          results <- rbind(results, rows) 
        }
        print("OTU") # logging
      },
      error = function(cond) {
        message("OTU-level TEST FAILED!")
        message(cond)
      }
    )
  
  # aggregate at each taxrank
  for (taxrank in rev(rank_names(ps)[-1])) {
    ps.taxa <- tax_glom(ps, taxrank = taxrank, NArm = TRUE)
    # skip if less than 10 taxa
    if (nrow(tax_table(ps.taxa)) < 11) {
      next
    }
    tryCatch(
      {
        # get result of test
        rows <- aldex2_on_growth_phase(ps.taxa, paste(test_id, taxrank, sep="_"))
        if (nrow(rows) > 0) {
          rows$taxrank <- taxrank
          results <- rbind(results, rows) 
        }
        print(taxrank) # logging
      },
      error = function(cond) {
        message("TEST FAILED!")
        message(cond)
      }
    )
  }
  
  if (nrow(results) > 0) {
    results$test_id <- test_id
    results$level <- "growth.phase" 
  }
  
  return(results)
}
```

## Cluster -- Pair-wise Comparisons

```{r}
set.seed(42)
combs <- combn(unique(sample_data(ps)$cluster), m=2, simplify=FALSE)
for (c in combs) {
  # get pair-wise combination
  g1 <- as.character(c[1])
  g2 <- as.character(c[2])
  test_id <- paste(substr(g1, 1, 1), substr(g2, 1, 1), sep="_vs_")
  # filter and run test
  print(test_id)
  rows <- drive_aldex2_on_cluster(
    phyloseq::subset_samples(ps, cluster == g1 | cluster == g2), 
    test_id)
  if (nrow(rows) > 0) {
    ALDEx2_results <- rbind(ALDEx2_results, rows)
  }
}
```

```{r}
ALDEx2_results
```

```{r}
unique(ALDEx2_results$test_id)
```

```{r}
unique(ALDEx2_results$level)
```

```{r}
ALDEx2_results_backup <- ALDEx2_results
```

## Focused Cluster -- Pair-wise Comparisons

```{r}
# rows <- drive_aldex2_on_growth_phase(
#           phyloseq::subset_samples(ps, growth.phase == "mid C" | growth.phase == "perturbed G"),
#           "test")
# rows
```

```{r}
set.seed(42)
combs <- combn(unique(sample_data(ps)$growth.phase), m=2, simplify=FALSE)
for (c in combs) {
  # get pair-wise combination
  g1 <- as.character(c[1])
  g2 <- as.character(c[2])
  test_id <- paste(g1, g2, sep=" vs ")
  # # filter and run test
  print(test_id)
  # get result of test
  rows <- drive_aldex2_on_growth_phase(
          phyloseq::subset_samples(ps, growth.phase == g1 | growth.phase == g2), 
          test_id)
  
  if (nrow(rows) > 0) {
    ALDEx2_results <- rbind(ALDEx2_results, rows)
  }
}
```

### Save to .csv

```{r}
nrow(ALDEx2_results_backup)
nrow(ALDEx2_results)
```
```{r}
ALDEx2_results
```


```{r}
unique(ALDEx2_results$test_id)
```

```{r}
unique(ALDEx2_results$level)
```

```{r}
# save to csv
write.csv(ALDEx2_results, paste(save_location, "ALDEx2_results.csv", sep=""), quote = FALSE)
```

```{r}
# save only significantly differentially abundant taxa
ALDEx2_results %>%
  filter(aldex2_wi.eBH < 0.05) %>%
  write.csv(paste(save_location, "ALDEx2_results_TRUE.csv", sep=""), quote = FALSE)
```

