---
title: "visualize differential abundance"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

Goal: since there is a significant difference in the composition of clusters, find out who is driving the difference

Groups
* "late" vs. "mid"

* "perturbed M" vs. "perturbed G"

* "mid C" vs. "mid G"
* "mid C" vs. "mid M"
* "mid G" vs. "mid M"

* "late C" vs. "late G" 
* "late C" vs. "late M"
* "late G" vs. "late M"

* "late C" vs. all the others

Methods:
* Wilcox Sum Rank Test
* ALDEx2
* ANCOM-BC

Levels
* Phylum
* Class
* Order
* Family
* Genus
* Species
* OTU-level

# Install Packages

```{r}
# .cran_packages <- c("tidyverse", "cowplot", "picante", "vegan", "HMP", "dendextend", "rms", "devtools")
# .inst <- .cran_packages %in% installed.packages()
# if(any(!.inst)) {
#   install.packages(.cran_packages[!.inst])
# }
```

```{r}
# .bioc_packages <- c("phyloseq", "Biostrings", "DESeq2", "microbiome", "metagenomeSeq", "ALDEx2")
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(.bioc_packages, version = "3.14")
```


# Loading Required Packages

```{r}
library("ggplot2"); packageVersion("ggplot2") # plotting
```
```{r}
library("tidyverse"); packageVersion("tidyverse") # util    
```

```{r}
library("dplyr"); packageVersion("dplyr") # util
```

```{r}
library("phyloseq"); packageVersion("phyloseq") # data object                 
```

```{r}
library("ape"); packageVersion("ape") # taxa string manipulation
```

```{r}
library(microbiome); packageVersion("microbiome") # clr transformation
```

```{r}
library(dendextend); packageVersion("dendextend") # extend customization of dendograms
```

```{r}
library(vegan); packageVersion("vegan") # bray-curtis and other metrics
```

```{r}
# Manipulation and visualization of taxonomic data, particularly those from amplicon metagenomics research. (Foster, Sharpton, and Grünwald 2017)
library(metacoder); packageVersion("metacoder")
```

```{r}
# defines taxonomic classes and functions to manipulate them. The goal is to use these classes as low level fundamental taxonomic classes that other R packages can build on and use. This is used by metacoder. 
library(taxa); packageVersion("taxa")
```

# Initialization

```{r}
# ggplot2 package theme set
theme_set(theme_void())

# control randomization
set.seed(42)

# save outputs
save_location = "../data/processed_carbon_53OTUs"
```

# Aes

```{r}
# color - days
color_day = c(
    "3" = "#B2E5EA", # lightest
    "5" = "#00BAE7",
    "7" = "#0E6FA0",
    "10" = "#012652",
    "20" = "#01091C"
    )

# color - growth.phase
color_growth.phase = c(
  "early" = "black",
  "mid C:C" = "springgreen3",
  "late C:C" = "springgreen4",
  "perturbed G" = "steelblue1",
  "mid G:C" = "steelblue3",
  "late G:C" = "steelblue4",
  "perturbed M" = "sienna1",
  "mid M:C" = "sienna3",
  "late M:C" = "sienna4"
  )

# shape - series
shape_series = c(
  "C0C" = 4,
  "C" = 16,
  "G" = 15,
  "M" = 17
  )
# shape - carbon
shape_carbon = c(
  "original" = 4,
  "cellulose" = 16,
  "glucose" = 18,
  "malate" = 15
)
# shape - transfer
shape_transfer = c(
    "C0C" = 4,
    "1C" = 10,
    "2C" = 16,
    "1G" = 0,
    "2G" = 12,
    "3G" = 15,
    "1M" = 5,
    "2M" = 9,
    "3M" = 18
  )
# shape - growth phase
shape_growth.phase = c(
  "early" = 1,
  "mid C:C" = 16,
  "late C:C" = 16,
  "perturbed G" = 15,
  "mid G:C" = 15,
  "late G:C" = 15,
  "perturbed M" = 17,
  "mid M:C" = 17,
  "late M:C" = 17
  )

# color - genus
```

# Load Phyloseq Data Object

```{r}
(ps <- readRDS(paste(save_location, "phyloseq.rds", sep="")))
# convert to taxmap format 
taxmap <- parse_phyloseq(ps)
taxmap
```

# Plot metacoder family tree

```{r}
set.seed(42)
heat_tree(taxmap,
          node_label = taxon_names,
          node_size = n_obs,
          node_color = n_obs,
          node_color_axis_label = "OTU count",
          layout = "da",
          initial_layout = "re",
          output_file = paste(save_location, "/metacoder_differential_abundance/carbon_all.png", sep = ""),
          title = "Carbon - All")
```

# Extract tax_ids

```{r}
taxmap.key <- merge(
  as.data.frame(taxon_names(taxmap)), 
  as.data.frame((taxon_ranks(taxmap))), 
  by.x=0, by.y=0)
colnames(taxmap.key) <- c("taxon_id", "taxon_names", "taxon_rank")
taxmap.key
```

```{r}
#Print number of duplicated rows 
#i.e. taxon not uniquely identified by taxon_name + taxon_rank
nrow(taxmap.key) - nrow(
  taxmap.key[!(duplicated(taxmap.key[c("taxon_names","taxon_rank")]) 
             | duplicated(taxmap.key[c("taxon_names","taxon_rank")],
            fromLast = TRUE)), ])
```

# Process DA .csv

```{r}
get_taxon_id = function(x) {
  #Get taxrank (finest resolution taxonomy)
  taxrank <- x[["taxrank"]]
  #Get tax name from relevant column
  taxname <- x[[taxrank]]
  
  #Query on tax_data
  if (taxrank == "OTU") {
    matches <- filter(taxmap$data$tax_data, otu_id == taxname)
    return(matches[1]$taxon_id)
  }
  
  #Query on taxmap.key
  matches <- filter(taxmap.key, taxon_names == taxname & taxon_rank == taxrank)
  return(matches[1]$taxon_id)
}
```

## ALDEx2

```{r}
#Read .csv
ALDEx2_results <- read.table(
  file = paste(save_location, '/ALDEx2_results.csv', sep=""), 
  header = TRUE, sep = ',')
#Get taxon_id
ALDEx2_results <- cbind(ALDEx2_results, taxon_id = apply(
  ALDEx2_results, 1, get_taxon_id))
```

Check which ALDEx2 has effect size > 1 but p-value > 0.05
```{r}
ALDEx2_results$aldex2_effect[ALDEx2_results$aldex2_wi.eBH > 0.05 & ALDEx2_results$aldex2_effect > 1]
```

Check which ALDEx2 has effect size < -1 but p-value > 0.05
```{r}
ALDEx2_results$aldex2_effect[ALDEx2_results$aldex2_wi.eBH > 0.05 & ALDEx2_results$aldex2_effect < -1]
```

```{r}
#Zero effect size for non-significant p-values
# ALDEx2_results$aldex2_effect[ALDEx2_results$aldex2_wi.eBH > 0.05] <- 0
#Show final table
ALDEx2_results
```

## Get Differential Abundance Table

```{r}
get_da_taxmap = function(taxmap, ti) {
  ALDEx2_results_filtered <- filter(ALDEx2_results, 
                                    test_id == ti & taxrank != "OTU")
  
  #Merge Dataframe
  taxmap$data$da <- taxmap.key %>%
    merge(ALDEx2_results_filtered, by.x = "taxon_id", by.y = "taxon_id", all.x = TRUE) %>%
    dplyr::select("taxon_id", "aldex2_diff.btw", "aldex2_diff.win", "aldex2_effect", "aldex2_positive", "aldex2_negative") %>%
    mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>%
    as.tibble()
  
  return(taxmap)
}
```

# Heat Map Trees

```{r}
# plot_comparative_heat_tree = function(ti) {
#   
#   #Get taxmap
#   tmp <- get_da_taxmap(taxmap, ti)
#   group1 <- unique(tmp$data$da$aldex2_positive)
#   group1 <- group1[!is.na(group1)][1];
#   group2 <- unique(tmp$data$da$aldex2_negative)
#   group2 <- group2[!is.na(group2)][1];
#   group1
#   group2
# 
#   #Get corresponding colors
#   blank_color <- "oldlace"
#   # blank_color <- "gray"
#   g1_color <- color_growth.phase[[group1]]
#   g2_color <- color_growth.phase[[group2]]
#   
#   #Get min, max
#   aldex2_min <- min(tmp$data$da$aldex2_effect)
#   aldex2_max <- max(tmp$data$da$aldex2_effect)
#   aldex2_range <- max(abs(aldex2_min), abs(aldex2_max))
#   
#   # ALDEx2
#   # control randomization
#   set.seed(42)
#   tmp %>%
#     filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
#     heat_tree(node_size = n_obs, # number of OTUs
#               node_size_axis_label = "OTU count",
#               node_color = aldex2_effect, # difference between groups
#               node_color_interval = c(-1 * aldex2_range, aldex2_range), # symmetric interval
#               node_color_range = c(g2_color, blank_color, g1_color), # diverging colors
#               node_color_axis_label = "ALDEx2 Effect Size",
#               # node_label = taxon_names, # comment to remove labels
#               initial_layout = "reingold-tilford",
#               layout = "davidson-harel",
#               title = paste(ti, "positive to", group1, "ALDEx2"),
#               output_file = paste(save_location,
#                 paste("metacoder_differential_abundance/heat_tree", 
#                 ti, "ALDEX2.png", sep="_"), sep=""))
#   
# }
```

```{r}
unique(ALDEx2_results$test_id)
```

```{r}
unique(ALDEx2_results$level)
```
 

```{r}
unique(c(ALDEx2_results$aldex2_positive, ALDEx2_results$aldex2_negative))
```

```{r}
# for (test_id in unique(ALDEx2_results$test_id)) {
#   tryCatch(
#     {
#       plot_comparative_heat_tree(test_id)
#       print(test_id)
#     },
#     error=function(cond) {
#       message(test_id)
#       message("-failed to run")
#     }
#   )  
# }
```

```{r}
## Get empty one with labels
# plot_comparative_heat_tree("mG_vs_mM")
```

# Scatter plots

```{r}
min(ALDEx2_results$aldex2_effect)
```

```{r}
max(ALDEx2_results$aldex2_effect)
```
```{r}
ALDEx2_results %>%
  ggplot(aes(x=aldex2_effect)) + 
  geom_histogram(binwidth=.5, colour="black", fill="white")

ALDEx2_results %>%
  filter(level == "cluster") %>%
  ggplot(aes(x=aldex2_effect)) + 
  geom_histogram(binwidth=.5, colour="black", fill="white")

ALDEx2_results %>%
  filter(level == "growth.phase") %>%
  ggplot(aes(x=aldex2_effect)) + 
  geom_histogram(binwidth=.5, colour="black", fill="white")
```
## Extra Aes

```{r}
color_growth.phase["not significant"] <- "gray50"

# for 20 color palette
tableau_20_colors <- ggthemes::ggthemes_data[["tableau"]][["color-palettes"]][["regular"]][["Tableau 20"]]$value
```


```{r}
get_da_scatter_ALDEx2 <- function(data, test_ids = c(), taxrank = "genus", plot_name="x") {
  # only use OTU level differential analysis
  # filter and color data points
  ALDEx2_results_filtered <- data %>%
    filter(test_id %in% test_ids) %>% 
    filter(taxrank == "OTU") %>% 
    # only plot significant OTUs
    # filter(aldex2_effect != 0) %>%
    # color scheme
    # mutate(significant = case_when(
    #   aldex2_wi.eBH < 0.05 & aldex2_effect > 1 ~ aldex2_positive,
    #   aldex2_wi.eBH < 0.05 & aldex2_effect < -1 ~ aldex2_negative,
    #   TRUE~"not significant"))
    mutate(significant = case_when(
      aldex2_effect > 1 ~ aldex2_positive,
      aldex2_effect < -1 ~ aldex2_negative,
      TRUE~"not significant"))
    
  # get number of y-axis categories
  num_categories <- length(unique(ALDEx2_results_filtered[[taxrank]]))
  taxrank_colors <- rep("black", num_categories)
  if (taxrank == "genus") {
    taxrank_colors <- c(tableau_20_colors, "black")
  }
  
  # get range of x-axis
  min_effect <- max(ALDEx2_results_filtered$aldex2_effect)
  max_effect <- min(ALDEx2_results_filtered$aldex2_effect)
  range_effect <- max(abs(min_effect), abs(max_effect))
  
  # divide taxrank categories into strips (for zebra striped background)
  strip_data <- data.frame(unique(ALDEx2_results_filtered[[taxrank]]))
  colnames(strip_data) <- c("taxrank")
  strip_data <- strip_data %>%
    arrange(desc(taxrank)) %>%
    mutate(xmin = -1 * range_effect, xmax = range_effect,
           y_position = rev(1:nrow(.)),
           ymin = y_position - 0.5,
           ymax = y_position + 0.5,
           fill = rep(c("zebra_a", "zebra_b"), length.out=nrow(strip_data)))
  
  # map taxrank to strip position
  refer_y_pos <- as.numeric(strip_data$y_position)
  names(refer_y_pos) <- strip_data$taxrank
  ALDEx2_results_filtered$y_position <- as.numeric(refer_y_pos[ALDEx2_results_filtered[[taxrank]]])
  ALDEx2_results_filtered
  
  # pivot strip data table
  strip_data <- strip_data %>%
    pivot_longer(cols=c(xmin, xmax), values_to="x", names_to="xmin_xmax") %>%
    dplyr::select(-xmin_xmax)
  
  # plot pretty
  ALDEx2_results_filtered %>%
    ggplot(axes = FALSE, show.legend = FALSE,
           aes_string(x = "aldex2_effect", 
                      y = "y_position", 
                      colour = "significant",
                      group="y_position")) +
    # zebra stripe background for readability
    geom_ribbon(data=strip_data,
                inherit.aes = FALSE, 
                show.legend = FALSE,
                aes(x=x, ymin=ymin, ymax=ymax, group=y_position, fill=fill)) +
    scale_fill_manual(name = 'fill', values = c("gray100", "gray95")) +
    scale_colour_manual(name = 'significant', values = color_growth.phase) +
    geom_jitter(width = 0, height = 0.10, alpha = 0.3) + 
    facet_grid(cols = vars(test_id)) + 
    scale_y_discrete(taxrank, limits=names(refer_y_pos)) +
    scale_x_continuous("Effect Size", breaks = c(-1, 0, 1)) + 
    geom_vline(xintercept=1, linetype="dotted", color="gray70", size=1) +
    geom_vline(xintercept=-1, linetype="dotted", color="gray70", size=1) +
    theme_set(theme_classic()) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    theme(strip.text.x = element_text(angle = 90)) +
    theme(legend.position = "none",
          panel.spacing = unit(0.1, "lines"),
          panel.border = element_rect(color = "black", fill = NA, size = 1.0), 
          strip.background = element_rect(color = "black", size = 1))
}
```

## Cluster level

```{r}
test_ids <- c("l_vs_e", "l_vs_m", "l_vs_p", "e_vs_m", "e_vs_p", "m_vs_p")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "genus")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "order")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "class")
```

## Focused Cluster level

```{r}
 unique(ALDEx2_results$test_id)
```
```{r}
test_ids <- c("late C:C vs early", "late C:C vs mid C:C", "late C:C vs perturbed M", "late C:C vs mid M:C", "late C:C vs late M:C", "late C:C vs perturbed G", "late C:C vs mid G:C", "late C:C vs late G:C", 
              "early vs mid C:C", "early vs perturbed M", "early vs mid M:C", "early vs late M:C", "early vs perturbed G", "early vs mid G:C", "early vs late G:C", 
              "mid C:C vs perturbed M", "mid C:C vs mid M:C", "mid C:C vs late M:C", "mid C:C vs mid G:C", "mid C:C vs late G:C", 
              "perturbed M vs mid M:C", "perturbed M vs perturbed G", "perturbed M vs mid G:C", "perturbed M vs late G:C",
              "mid M:C vs late M:C", "mid M:C vs perturbed G", "mid M:C vs mid G:C", "mid M:C vs late G:C", 
              "late M:C vs perturbed G", "late M:C vs mid G:C", "late M:C vs late G:C", 
              "perturbed G vs mid G:C", "perturbed G vs late G:C",
              "mid G:C vs late G:C")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "genus", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "order", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "class", plot_name=plot_name)
```

```{r}
test_ids <- c("late C:C vs early", "late C:C vs mid C:C", "late C:C vs perturbed M", "late C:C vs mid M:C", "late C:C vs late M:C", "late C:C vs perturbed G", "late C:C vs mid G:C", "late C:C vs late G:C", 
              "mid C:C vs perturbed M", "mid C:C vs mid M:C", "mid C:C vs late M:C", "mid C:C vs mid G:C", "mid C:C vs late G:C")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "genus", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "order", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "class", plot_name=plot_name)
```

```{r}
test_ids <- c("early vs mid C:C", "early vs perturbed M", "early vs mid M:C", "early vs late M:C", "early vs perturbed G", "early vs mid G:C", "early vs late G:C")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "genus", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "order", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "class", plot_name=plot_name)
```

```{r}
test_ids <- c("perturbed M vs mid M:C", "perturbed M vs perturbed G", "perturbed M vs mid G:C", "perturbed M vs late G:C",
              "mid M:C vs late M:C", "mid M:C vs perturbed G", "mid M:C vs mid G:C", "mid M:C vs late G:C", 
              "late M:C vs perturbed G", "late M:C vs mid G:C", "late M:C vs late G:C", 
              "perturbed G vs mid G:C", "perturbed G vs late G:C",
              "mid G:C vs late G:C")
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "genus", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "order", plot_name=plot_name)
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = test_ids, taxrank = "class", plot_name=plot_name)
```







