---
title: "Differential Abundance Focus"
output: html_notebook
    fig_width: 16
    fig_height: 16
author: Gati Aher
date: 12/07/2021
fig_width: 24  
fig_height: 16
---

# Differential Abundance Testing

Goal: identify specific taxa associated with clinical metadata variables of interest
* difficult
* controversial
* concerns that normalization and testing approaches have generally failed to control false discovery rates
* contributed to lack of reproducibility in microbiome studies

Two Common Approaches
1. Applying non-parametric Wilcoxon rank-sum test to each taxon
2. Version of Wilcoxon test developed for compositional NGS data

Include the use of a CoDA transform for both since there does seem to be some growing support that log-ratio methodologies may better control the false positive rate.

```{r}
.cran_packages <- c("tidyverse", "cowplot", "picante", "vegan", "HMP", "dendextend", "rms", "devtools")
```

```{r}
.bioc_packages <- c("phyloseq", "DESeq2", "microbiome", "metagenomeSeq", "ALDEx2")
```

```{r}
.inst <- .cran_packages %in% installed.packages()
if(any(!.inst)) {
  install.packages(.cran_packages[!.inst])
}
```

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(.bioc_packages, version = "3.14")
```

# Loading Required Packages

```{r}
# Popular package with tools for analysis and visualization of microbiome data (McMurdie and Holmes 2013).
library(phyloseq); packageVersion("phyloseq")                 
```

```{r}
# Implements many standard statistical techniques used in ecological research. Is used extensively by packages like phyloseq and metacoder. (Dixon 2003)
library(vegan); packageVersion("vegan")
```

```{r}
# Manipulation and visualization of taxonomic data, particularly those from amplicon metagenomics research. (Foster, Sharpton, and Grünwald 2017)
library(metacoder); packageVersion("metacoder")
```

```{r}
# defines taxonomic classes and functions to manipulate them. The goal is to use these classes as low level fundamental taxonomic classes that other R packages can build on and use. This is used by metacoder. 
library(taxa); packageVersion("taxa")
```

```{r}
# Awesome graphing package. (Wickham 2009)
library(ggplot2); packageVersion("ggplot2")
```

```{r}
# A package for manipulating tabular data with a cohesive and intuitive set of commands. A popular alternative to base R methods.
library(dplyr); packageVersion("dplyr")
```

```{r}
# Makes reading tabular data from files easier.
library(readr); packageVersion("readr")
```

```{r}
# Functions for text manipulation.
library(stringr); packageVersion("stringr")
```

```{r}
# Used for the design and analysis of experiments, especially plant-related experiments.
library(agricolae); packageVersion("agricolae")
```

```{r}
# A popular package for DNA sequence analysis and phylogenetics.
library(ape); packageVersion("ape")
```
```{r}
# has clr transform
library(microbiome); packageVersion("microbiome")
```


# Load Processed Data Into R

```{r}
ps <- readRDS("../data/r/phyloseq.rds")
ps
```

```{r}
# use only the clr transformed data
(ps <- microbiome::transform(ps, "clr"))
```

# Create Taxmap Format -- Used By Metacoder

```{r}
(taxmap <- parse_phyloseq(ps))
```

```{r}
set.seed(0)
taxmap %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs,
            node_color = n_obs,
            node_color_axis_label = "Log-Ratio",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel")
```
# Check Differential Abundance of Samples day 5-20, transfer 1

```{r}
# create subset sample metadata
meta_df <- data.frame(sample_data(ps))
meta_df <- subset(meta_df, day %in% c(5, 7, 10, 20))
meta_df <- subset(meta_df, transfer == 1)
row.names(meta_df)
```

```{r}
(taxmap$data$carbon_abund <- calc_taxon_abund(taxmap, "otu_table", 
                                              cols = row.names(meta_df), 
                                              groups = meta_df$carbon))
```

```{r}
taxmap$data$tax_abund <- calc_taxon_abund(taxmap, "otu_table", 
                                           cols = row.names(meta_df))

taxmap$data$diff_table <- compare_groups(taxmap, dataset = "tax_abund",
                                         cols = row.names(meta_df),
                                         groups = meta_df$carbon)
  
taxmap <- mutate_obs(taxmap, "diff_table",
                  wilcox_p_value = p.adjust(wilcox_p_value, method = "fdr"))
  
taxmap$data$diff_table$log2_median_ratio[taxmap$data$diff_table$wilcox_p_value > 0.05] <- 0
```
```{r}
print(taxmap$data$diff_table)
```
```{r}
range(taxmap$data$diff_table$wilcox_p_value, finite = TRUE) 
```

```{r}
# plot heat tree
set.seed(0)
taxmap %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  # filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  heat_tree_matrix(dataset = "diff_table",
                   node_label = taxon_names,
                   node_size = n_obs, # number of OTUs
                   node_color = log2_median_ratio, # difference between groups
                   node_color_trans = "linear",
                   node_color_interval = c(-3, 3), # symmetric interval
                   edge_color_interval = c(-3, 3), # symmetric interval
                   node_color_range = diverging_palette(), # diverging colors
                   node_size_axis_label = "Log-Ratio",
                   node_color_axis_label = "Log 2 ratio of median counts",
                   layout = "da", 
                   initial_layout = "re",
                   output_file = "../terminal/heat_tree_day_5_to_20.pdf")
```
# Check Differential Abundance of Samples on/returned to cellulose, day 5-20

```{r}
# create subset sample metadata
meta_df_cellulose <- data.frame(sample_data(ps))
meta_df_cellulose <- subset(meta_df_cellulose, carbon == "cellulose")
meta_df_cellulose <- subset(meta_df_cellulose, day %in% c(5, 7, 10, 20))

# sort row by stability (order: "transitioning", "stable", "early")
# meta_df_cellulose$stability <- apply(meta_df_cellulose, 1, FUN = function(x) if(x["day"] == 3) "c" else if ((x["day"] == 5) | (x["day"] == 7)) "a" else "b")
# meta_df_cellulose <- meta_df_cellulose[order(meta_df_cellulose$stability),]
# meta_df_cellulose$stability <- apply(meta_df_cellulose, 1, FUN = function(x) if(x["day"] == 3) "early" else if ((x["day"] == 5) | (x["day"] == 7)) "transitioning" else "stable")
row.names(meta_df_cellulose)
```

```{r}
taxmap$data$tax_abund <- calc_taxon_abund(taxmap, "otu_table", 
                                           cols = row.names(meta_df_cellulose))

taxmap$data$diff_table <- compare_groups(taxmap, dataset = "tax_abund",
                                         cols = row.names(meta_df_cellulose),
                                         groups = meta_df_cellulose$group)
  
taxmap <- mutate_obs(taxmap, "diff_table",
                  wilcox_p_value = p.adjust(wilcox_p_value, method = "fdr"))
  
taxmap$data$diff_table$log2_median_ratio[taxmap$data$diff_table$wilcox_p_value > 0.05] <- 0
```

```{r}
print(taxmap$data$diff_table)

range(taxmap$data$diff_table$wilcox_p_value, finite = TRUE)

# plot heat tree
set.seed(0)
taxmap %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  # filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  heat_tree_matrix(dataset = "diff_table",
                   node_label = taxon_names,
                   node_size = n_obs, # number of OTUs
                   node_color = log2_median_ratio, # difference between groups
                   node_color_trans = "linear",
                   node_color_interval = c(-3, 3), # symmetric interval
                   edge_color_interval = c(-3, 3), # symmetric interval
                   node_color_range = diverging_palette(), # diverging colors
                   node_size_axis_label = "Log-Ratio",
                   node_color_axis_label = "Log 2 ratio of median counts",
                   layout = "da", 
                   initial_layout = "re",
                   output_file = "../terminal/heat_tree_only_cellulose_day_5_to_20.pdf"
                   )
```

# Check Differential Abundance of Samples on/returned to cellulose, day 5-20, do not show taxa that are not in subsampled set

```{r}
ps_cellulose <- phyloseq::subset_samples(ps, carbon =="cellulose")
(ps_cellulose <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_cellulose) > 0, ps_cellulose))
```

```{r}
(taxmap_cellulose <- parse_phyloseq(ps_cellulose))
```

```{r}
# create subset sample metadata
meta_df_cellulose <- data.frame(sample_data(ps_cellulose))
meta_df_cellulose <- subset(meta_df_cellulose, carbon == "cellulose")
meta_df_cellulose <- subset(meta_df_cellulose, day %in% c(5, 7, 10, 20))

# sort row by stability (order: "transitioning", "stable", "early")
# meta_df_cellulose$stability <- apply(meta_df_cellulose, 1, FUN = function(x) if(x["day"] == 3) "c" else if ((x["day"] == 5) | (x["day"] == 7)) "a" else "b")
# meta_df_cellulose <- meta_df_cellulose[order(meta_df_cellulose$stability),]
# meta_df_cellulose$stability <- apply(meta_df_cellulose, 1, FUN = function(x) if(x["day"] == 3) "early" else if ((x["day"] == 5) | (x["day"] == 7)) "transitioning" else "stable")
row.names(meta_df_cellulose)
```

```{r}
taxmap_cellulose$data$tax_abund <- calc_taxon_abund(taxmap_cellulose, "otu_table", 
                                           cols = row.names(meta_df_cellulose))

taxmap_cellulose$data$diff_table <- compare_groups(taxmap_cellulose, dataset = "tax_abund",
                                         cols = row.names(meta_df_cellulose),
                                         groups = meta_df_cellulose$group)
  
taxmap_cellulose <- mutate_obs(taxmap_cellulose, "diff_table",
                  wilcox_p_value = p.adjust(wilcox_p_value, method = "fdr"))
  
taxmap_cellulose$data$diff_table$log2_median_ratio[taxmap_cellulose$data$diff_table$wilcox_p_value > 0.05] <- 0
```

```{r}
print(taxmap_cellulose$data$diff_table)

range(taxmap_cellulose$data$diff_table$wilcox_p_value, finite = TRUE)

# plot heat tree
set.seed(0)
taxmap_cellulose %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  # filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  heat_tree_matrix(dataset = "diff_table",
                   node_label = taxon_names,
                   node_size = n_obs, # number of OTUs
                   node_color = log2_median_ratio, # difference between groups
                   node_color_trans = "linear",
                   node_color_interval = c(-3, 3), # symmetric interval
                   edge_color_interval = c(-3, 3), # symmetric interval
                   node_color_range = diverging_palette(), # diverging colors
                   node_size_axis_label = "Log-Ratio",
                   node_color_axis_label = "Log 2 ratio of median counts",
                   layout = "da", 
                   initial_layout = "re",
                   output_file = "../terminal/heat_tree_filtered_taxa_only_cellulose_day_5_to_20.pdf"
                   )
```