---
title: "differential abundance metacoder"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

Goal: since there is a significant difference in the composition of clusters, find out who is driving the difference

Groups
* "stable" vs. "recovering"

* "perturbed M" vs. "perturbed G"

* "recovering C" vs. "recovering G"
* "recovering C" vs. "recovering M"
* "recovering G" vs. "recovering M"

* "stable C" vs. "stable G" 
* "stable C" vs. "stable M"
* "stable G" vs. "stable M"

Methods:
* Wilcox Sum Rank Test
* ALDEx2
* ANCOM-BC

Levels
* Phylum
* Class
* Order
* Family
* Genus
* Species
* OTU-level

# Install Packages

```{r}
# .cran_packages <- c("tidyverse", "cowplot", "picante", "vegan", "HMP", "dendextend", "rms", "devtools")
# .inst <- .cran_packages %in% installed.packages()
# if(any(!.inst)) {
#   install.packages(.cran_packages[!.inst])
# }
```

```{r}
# .bioc_packages <- c("phyloseq", "Biostrings", "DESeq2", "microbiome", "metagenomeSeq", "ALDEx2")
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(.bioc_packages, version = "3.14")
```

# Loading Required Packages

```{r}
library("ggplot2"); packageVersion("ggplot2") # plotting
```

```{r}
library("tidyverse"); packageVersion("tidyverse") # util              
```

```{r}
library("dplyr"); packageVersion("dplyr") # util
```

```{r}
library("phyloseq"); packageVersion("phyloseq") # data object                 
```

```{r}
library("ape"); packageVersion("ape") # taxa string manipulation
```

```{r}
library(microbiome); packageVersion("microbiome") # clr transformation
```

```{r}
library(dendextend); packageVersion("dendextend") # extend customization of dendograms
```

```{r}
library(vegan); packageVersion("vegan") # bray-curtis and other metrics
```

```{r}
# Manipulation and visualization of taxonomic data, particularly those from amplicon metagenomics research. (Foster, Sharpton, and Grünwald 2017)
library(metacoder); packageVersion("metacoder")
```

```{r}
# defines taxonomic classes and functions to manipulate them. The goal is to use these classes as low level fundamental taxonomic classes that other R packages can build on and use. This is used by metacoder. 
library(taxa); packageVersion("taxa")
```

# Initialization

```{r}
# ggplot2 package theme set
theme_set(theme_void())

# control randomization
set.seed(42)

# save outputs
save_location = "../data/processed_FCF/"
```

# Load Phyloseq Data Object

```{r}
(ps <- readRDS(paste(save_location, "phyloseq.rds", sep="")))
# convert to taxmap format 
taxmap <- parse_phyloseq(ps)
taxmap
```

# Extract tax_ids

```{r}
taxmap.key <- merge(
  as.data.frame(taxon_names(taxmap)), 
  as.data.frame((taxon_ranks(taxmap))), 
  by.x=0, by.y=0)
colnames(taxmap.key) <- c("taxon_id", "taxon_names", "taxon_rank")
taxmap.key
```

```{r}
#Print number of duplicated rows 
#i.e. taxon not uniquely identified by taxon_name + taxon_rank
nrow(taxmap.key) - nrow(
  taxmap.key[!(duplicated(taxmap.key[c("taxon_names","taxon_rank")]) 
             | duplicated(taxmap.key[c("taxon_names","taxon_rank")],
            fromLast = TRUE)), ])
```

# Process DA .csv

```{r}
get_taxon_id = function(x) {
  #Get taxrank (finest resolution taxonomy)
  taxrank <- x[["taxrank"]]
  #Get tax name from relevant column
  taxname <- x[[taxrank]]
  
  #Query on tax_data
  if (taxrank == "OTU") {
    matches <- filter(taxmap$data$tax_data, otu_id == taxname)
    return(matches[1]$taxon_id)
  }
  
  #Query on taxmap.key
  matches <- filter(taxmap.key, taxon_names == taxname & taxon_rank == taxrank)
  return(matches[1]$taxon_id)
}
```

## ANCOM-BC

```{r}
#Read .csv
ANCOM_results <- read.table(
  file = paste(save_location, 'ANCOM_results.csv', sep=""),
  header = TRUE, sep = ',')
#Get taxon_id
ANCOM_results <- cbind(ANCOM_results, taxon_id = apply(
  ANCOM_results, 1, get_taxon_id))
#Zero effect size for non-significant p-values
ANCOM_results$ancom_w[ANCOM_results$ancom_q_val > 0.05] <- 0
#Show final table
ANCOM_results
```

```{r}
unique(ANCOM_results$test_id)
```

## ALDEx2

```{r}
#Read .csv
ALDEx2_results <- read.table(
  file = paste(save_location, 'ALDEx2_results.csv', sep=""), 
  header = TRUE, sep = ',')
#Get taxon_id
ALDEx2_results <- cbind(ALDEx2_results, taxon_id = apply(
  ALDEx2_results, 1, get_taxon_id))
#Zero effect size for non-significant p-values
ALDEx2_results$aldex2_effect[ALDEx2_results$aldex2_wi.eBH > 0.05] <- 0
#Show final table
ALDEx2_results
```

## Get Differential Abundance Table

```{r}
get_da_taxmap = function(taxmap, ti, group1, group2) {
  ANCOM_results_filtered <- filter(ANCOM_results, 
                                   test_id == ti & taxrank != "OTU")
  ALDEx2_results_filtered <- filter(ALDEx2_results, 
                                    test_id == ti & taxrank != "OTU")
  
  #If test magnitude is relative to different group than group 1
  #then negate effect size variable
  if(nrow(ANCOM_results_filtered) > 0 &
     ANCOM_results_filtered$ancom_relative_to[1] != group1) {
      # invert relative_to label
      ANCOM_results_filtered$ancom_relative_to <- group1
      # invert results
      ANCOM_results_filtered$ancom_W <- ANCOM_results_filtered$ancom_W * -1
  }
  if(nrow(ALDEx2_results_filtered) > 0 &
     ALDEx2_results_filtered$aldex2_relative_to[1] != group1) {
      # invert relative_to label
      ALDEx2_results_filtered$aldex2_relative_to <- group1
      # invert results
      ALDEx2_results_filtered$aldex2_effect <- ALDEx2_results_filtered$aldex2_effect * -1
  }
  
  #Merge Dataframe
  taxmap$data$da <- taxmap.key %>%
    merge(ANCOM_results_filtered, by.x = "taxon_id", by.y = "taxon_id", all.x = TRUE) %>%
    merge(ALDEx2_results_filtered, by.x = "taxon_id", by.y = "taxon_id", all.x = TRUE) %>%
    select("taxon_id", "aldex2_diff.btw", "aldex2_diff.win", "aldex2_effect", "ancom_relative_to", "ancom_W") %>%
    mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>%
    as.tibble()
  
  return(taxmap)
}
```

# Heat Map Trees

```{r}
plot_comparative_heat_tree = function(ti, group1, group2) {
  
  #Define color key
  focused.cluster_colors <- c(
                          "black", "slateblue1", "sienna1",
                          "springgreen3", "slateblue3", "sienna3", 
                          "springgreen4", "slateblue4", "sienna4",
                          "red", "blue")
  names(focused.cluster_colors) <- c(
                          "early", "perturbed G", "perturbed M",
                          "recovering C", "recovering G", "recovering M",
                          "stable C", "stable G", "stable M",
                          "recovering", "stable")
  
  #Get corresponding colors
  blank_color <- "oldlace"
  # blank_color <- "gray"
  g1_color <- focused.cluster_colors[[group1]]
  g2_color <- focused.cluster_colors[[group2]]
  
  #Get taxmap
  tmp <- get_da_taxmap(taxmap, ti, group1, group2)
  
  #Get min, max
  ancom_min <- min(tmp$data$da$ancom_W)
  ancom_max <- max(tmp$data$da$ancom_W)
  ancom_range <- max(abs(ancom_min), abs(ancom_max))
  
  aldex2_min <- min(tmp$data$da$aldex2_effect)
  aldex2_max <- max(tmp$data$da$aldex2_effect)
  aldex2_range <- max(abs(aldex2_min), abs(aldex2_max))
  
  # ANCOM-BC
  # control randomization
  set.seed(42)
  tmp %>%
    filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
    heat_tree(node_size = n_obs, # number of OTUs
              node_size_axis_label = "OTU count",
              node_color = ancom_W, # difference between groups
              node_color_interval = c(-1 * ancom_range, ancom_range), # symmetric interval
              node_color_range = c(g2_color, blank_color, g1_color), # diverging colors
              node_color_axis_label = "ANCOM Effect Size",
              # node_label = taxon_names, # comment to remove labels
              initial_layout = "reingold-tilford",
              layout = "davidson-harel",
              title = paste(ti, "positive to", group1, "ANCOM"),
              output_file = paste(save_location,
                paste("metacoder_differential_abundance/heat_tree", 
                ti, "ANCOM.png", sep="_"), sep=""))
  
  # ALDEx2
  # control randomization
  set.seed(42)
  tmp %>%
    filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
    heat_tree(node_size = n_obs, # number of OTUs
              node_size_axis_label = "OTU count",
              node_color = aldex2_effect, # difference between groups
              node_color_interval = c(-1 * aldex2_range, aldex2_range), # symmetric interval
              node_color_range = c(g2_color, blank_color, g1_color), # diverging colors
              node_color_axis_label = "ALDEx2 Effect Size",
              # node_label = taxon_names, # comment to remove labels
              initial_layout = "reingold-tilford",
              layout = "davidson-harel",
              title = paste(ti, "positive to", group1, "ALDEx2"),
              output_file = paste(save_location,
                paste("metacoder_differential_abundance/heat_tree", 
                ti, "ALDEX2.png", sep="_"), sep=""))
  
}
```

```{r}
unique(ALDEx2_results$test_id)
```

```{r}
unique(ALDEx2_results$aldex2_relative_to)
```

```{r}
plot_comparative_heat_tree("early_vs_recovering", "early", "recovering")
```
```{r}
plot_comparative_heat_tree("early_vs_stable", "early", "stable")
```

```{r}
plot_comparative_heat_tree("stable_vs_recovering", "stable", "recovering")
```

```{r}
plot_comparative_heat_tree("pG_vs_pM", "perturbed G", "perturbed M")
```

```{r}
plot_comparative_heat_tree("rC_vs_rM", "recovering C", "recovering M")
```

```{r}
plot_comparative_heat_tree("rC_vs_rG", "recovering C", "recovering G")
```

```{r}
plot_comparative_heat_tree("rG_vs_rM", "recovering G", "recovering M")
```

```{r}
plot_comparative_heat_tree("sC_vs_sG", "stable C", "stable G")
```

```{r}
plot_comparative_heat_tree("sG_vs_sM", "stable G", "stable M")
```

```{r}
plot_comparative_heat_tree("sC_vs_sM", "stable C", "stable M")
```

## Focused Cluster -- Pair-wise Comparisons to sC

```{r}
plot_comparative_heat_tree("sC_vs_pG", "stable C", "perturbed G")
```
```{r}
# empty, using as key, don't overwrite
# plot_comparative_heat_tree("sC_vs_pM", "stable C", "perturbed M")
```

```{r}
plot_comparative_heat_tree("sC_vs_rC", "stable C", "recovering C")
```
```{r}
plot_comparative_heat_tree("sC_vs_rG", "stable C", "recovering G")
```
```{r}
plot_comparative_heat_tree("sC_vs_rM", "stable C", "recovering M")
```

## Get one with labels
```{r}
# empty
# plot_comparative_heat_tree("sC_vs_pM", "stable C", "perturbed M")
```

# Scatter plots

```{r}
# ggplot2 package theme set
theme_set(theme_gray())
```

```{r}
unique(ALDEx2_results$test_id)
```

```{r}
unique(ALDEx2_results$aldex2_relative_to)
```

```{r}

```


```{r}
get_da_scatter_ALDEx2 <- function(data, test_ids = c(), taxrank = "genus") {
  # only use OTU level differential analysis
  # flag OTU that are significant
  ALDEx2_results_filtered <- data %>%
    filter(test_id %in% test_ids) %>% 
    filter(taxrank == "OTU") %>% 
    mutate(significant = case_when(
      aldex2_wi.eBH < 0.05 & aldex2_effect < 0 ~ "other group",
      aldex2_wi.eBH < 0.05 & aldex2_effect > 0 ~ aldex2_relative_to,
      TRUE~"not significant"
  ))
  
  ggplot(data = ALDEx2_results_filtered) +
  aes_string(x = "aldex2_effect", 
      y = taxrank,
      colour = "significant") + 
  scale_colour_manual(name = 'significant', 
                      values = setNames(c("black", "slateblue1", "sienna1",
                                          "springgreen3", "slateblue3", "sienna3", 
                                          "springgreen4", "slateblue4", "sienna4",
                                          "red", "blue", "firebrick"),
                                        c("early", "perturbed G", "perturbed M",
                                          "recovering C", "recovering G", "recovering M",
                                          "stable C", "stable G", "stable M",
                                          "recovering", "stable", "other group")
                          )) + 
  geom_jitter() + 
  facet_grid(cols = vars(test_id))
    
}
```


```{r}
get_da_scatter_ALDEx2(data = ALDEx2_results, test_ids = c("early_vs_recovering", "early_vs_stable", "stable_vs_recovering"), taxrank = "genus")
```




