---
title: "Using robCompositions -- Failure"
output: html_notebook
---

Goal: Make PCA on IRL Data and Biplot on CLR Data

Resources:
* [Exploratory compositional data analysis using the
R-package robCompositions (Hron, Templ, Filzmoser)](http://file.statistik.tuwien.ac.at/filz/papers/2010CDAM2.pdf)

ISSUE: [pcaCODA doesn't work with microbiome data](https://github.com/matthias-da/robCompositions/issues/10)


```{r}
# install.packages("devtools")
# library(devtools)
# install_github("matthias-da/robCompositions")

# install.packages("robCompositions")
library(robCompositions)
```

## Load Compositional Data

Data Format
* sample x OTU counts

```{r}
data("expenditures")
```

## Compute Robust Mahalanobis Distance Matrix

* internally applies isometric log-ratio transformation to the compositions to search for outliers in the real space

```{r}
# check for outliers (same detection results when method="standard")
outRob <- outCoDa(expenditures, method="robust")
outRob
```
```{r}
outRob$mahalDist
```
```{r}
outRob$outlierIndex
```
```{r}
# Figure 1: Graphical output by outlier detection for the expenditures data set.
plot(outRob)
```
 
```{r}
# robust compositional biplot
PrinCompRob <- pcaCoDa(expenditures, method="robust", solve="eigen")
PrinCompRob
summary(PrinCompRob)
plot(PrinCompRob)
biplot(PrinCompRob)
```
 
```{r}
# original data set expressed in percentages
constSum(expenditures)
```
 
