---
title: "differential abundance wilcoxon sum rank"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

Goal: since there is a significant difference in the composition of
clusters, find out who is driving the difference

Groups \* "stable" vs. "recovering" \* "perturbed M" vs. "perturbed G"

-   "recovering"

-   "recovering C" vs. "recovering G"

-   "recovering C" vs. "recovering M"

-   "recovering G" vs. "recovering M"

-   "stable"

-   "stable C" vs. "stable G"

-   "stable C" vs. "stable M"

-   "stable G" vs. "stable M"

Methods: \* Wilcox Sum Rank Test \* ALDEx2 \* ANCOM-BC

Levels \* Phylum \* Class \* Order \* Family \* Genus \* Species

# Install Packages

```{r}
# .cran_packages <- c("tidyverse", "cowplot", "picante", "vegan", "HMP", "dendextend", "rms", "devtools")
# .inst <- .cran_packages %in% installed.packages()
# if(any(!.inst)) {
#   install.packages(.cran_packages[!.inst])
# }
```

```{r}
# .bioc_packages <- c("phyloseq", "Biostrings", "DESeq2", "microbiome", "metagenomeSeq", "ALDEx2")
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(.bioc_packages, version = "3.14")
```

# Loading Required Packages

```{r}
library("ggplot2"); packageVersion("ggplot2") # plotting
```

```{r}
library("tidyverse"); packageVersion("tidyverse") # util              
```

```{r}
library("dplyr"); packageVersion("dplyr") # util
```

```{r}
library("phyloseq"); packageVersion("phyloseq") # data object                 
```

```{r}
library("ape"); packageVersion("ape") # taxa string manipulation
```

```{r}
library(microbiome); packageVersion("microbiome") # clr transformation
```

```{r}
library(dendextend); packageVersion("dendextend") # extend customization of dendograms
```

```{r}
library(vegan); packageVersion("vegan") # bray-curtis and other metrics
```

```{r}
# Manipulation and visualization of taxonomic data, particularly those from amplicon metagenomics research. (Foster, Sharpton, and Grünwald 2017)
library(metacoder); packageVersion("metacoder")
```

```{r}
# defines taxonomic classes and functions to manipulate them. The goal is to use these classes as low level fundamental taxonomic classes that other R packages can build on and use. This is used by metacoder. 
library(taxa); packageVersion("taxa")
```

# Initialization

```{r}
# ggplot2 package theme set
theme_set(theme_bw())

# control randomization
set.seed(42)
```

# Load Phyloseq Data Object

```{r}
(ps <- readRDS("../data/processed/phyloseq.rds"))
# convert to taxmap format 
taxmap <- parse_phyloseq(ps)

# Get Relative Abundance transform values
ps_rel_abund = phyloseq::transform_sample_counts(ps, function(x){x / sum(x)})

# Get CLR transform values
ps_clr <- microbiome::transform(ps, "clr")
taxmap_clr <- parse_phyloseq(ps_clr)
```

```{r}
taxon_names(taxmap)
```


```{r}
taxon_ranks(taxmap)
```

# Make Subset Cluster Group Tables

-   "stable" vs. "recovering"

```{r}
(ps_stable_vs_recovering <- phyloseq::subset_samples(ps, cluster == "stable" | cluster == "recovering"))
(ps_stable_vs_recovering <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stable_vs_recovering) > 0, ps_stable_vs_recovering))
```

-   "perturbed M" vs. "perturbed G"

```{r}
(ps_perturbedM_vs_perturbedG <- phyloseq::subset_samples(ps, focused.cluster == "perturbed M" | focused.cluster == "perturbed G"))
(ps_perturbedM_vs_perturbedG <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_perturbedM_vs_perturbedG) > 0, ps_perturbedM_vs_perturbedG))
```

-   "recovering"
-   "recovering C" vs. "recovering G"
-   "recovering C" vs. "recovering M"
-   "recovering G" vs. "recovering M"

```{r}
(ps_recovering <- phyloseq::subset_samples(ps, cluster == "recovering"))
(ps_recovering <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_recovering) > 0, ps_recovering))

(ps_recoveringC_vs_recoveringG <- phyloseq::subset_samples(ps, focused.cluster == "recovering C" | focused.cluster == "recovering G"))
(ps_recoveringC_vs_recoveringG <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_recoveringC_vs_recoveringG) > 0, ps_recoveringC_vs_recoveringG))

(ps_recoveringC_vs_recoveringM <- phyloseq::subset_samples(ps, focused.cluster == "recovering C" | focused.cluster == "recovering M"))
(ps_recoveringC_vs_recoveringM <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_recoveringC_vs_recoveringM) > 0, ps_recoveringC_vs_recoveringM))

(ps_recoveringG_vs_recoveringM <- phyloseq::subset_samples(ps, focused.cluster == "recovering G" | focused.cluster == "recovering M"))
(ps_recoveringG_vs_recoveringM <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_recoveringG_vs_recoveringM) > 0, ps_recoveringG_vs_recoveringM))
```

-   "stable"
-   "stable C" vs. "stable G",
-   "stable C" vs. "stable M"
-   "stable G" vs. "stable M"

```{r}
(ps_stable <- phyloseq::subset_samples(ps, cluster == "stable"))
(ps_stable <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stable) > 0, ps_stable))

(ps_stableC_vs_stableG <- phyloseq::subset_samples(ps, focused.cluster == "stable C" | focused.cluster == "stable G"))
(ps_stableC_vs_stableG <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stableC_vs_stableG) > 0, ps_stableC_vs_stableG))

(ps_stableC_vs_stableM <- phyloseq::subset_samples(ps, focused.cluster == "stable C" | focused.cluster == "stable M"))
(ps_stableC_vs_stableM <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stableC_vs_stableM) > 0, ps_stableC_vs_stableM))

(ps_stableG_vs_stableM <- phyloseq::subset_samples(ps, focused.cluster == "stable G" | focused.cluster == "stable M"))
(ps_stableG_vs_stableM <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stableG_vs_stableM) > 0, ps_stableG_vs_stableM))
```

# Wilcox Sum Rank Test

```{r}
save_diff_abund_tax <- function(taxmap, save_to) {
  sig_taxmap <- filter_obs(taxmap, data = "diff_table", drop_taxa = TRUE, wilcox_p_value < 0.05)
  
  sig_wilcoxon <- merge(x=sig_taxmap$data$diff_table, y=sig_taxmap$data$tax_data, by="taxon_id", all.x=TRUE) %>%
    arrange(log2_median_ratio, wilcox_p_value)
  
  write.csv(sig_wilcoxon, save_to, quote = FALSE)
  
  return(sig_wilcoxon)
}


wilcoxon_rank_sum_cluster <- function(ps, save_name) {
  # convert to taxmap object
  taxmap <- parse_phyloseq(ps)

  # get meta df
  meta_df <- data.frame(sample_data(ps))
  row.names(meta_df)
  
  # calculate taxon abundance
  taxmap$data$taxon_abund <- calc_taxon_abund(taxmap, "otu_table", cols = row.names(meta_df))

  # calculate wilcoxon rank sum test
  taxmap$data$diff_table <- compare_groups(taxmap, dataset = "taxon_abund",
                                         cols = row.names(meta_df),
                                         groups = meta_df$cluster)

  # adjust p-values
  taxmap <- mutate_obs(taxmap, "diff_table",
                  wilcox_p_value = p.adjust(wilcox_p_value, method = "fdr"))

  # adjust value to zero if not significant p-value
  taxmap$data$diff_table$log2_median_ratio[taxmap$data$diff_table$wilcox_p_value > 0.05] <- 0
  
  # save significant names and values to csv
  save_diff_abund_tax(taxmap, paste("../data/metacoder_differential_abundance/", save_name, ".csv", sep=""))

  return(taxmap)
}

wilcoxon_rank_sum_focused_cluster <- function(ps, save_name) {
  # convert to taxmap object
  taxmap <- parse_phyloseq(ps)

  # get meta df
  meta_df <- data.frame(sample_data(ps))
  row.names(meta_df)
  
  # calculate taxon abundance
  taxmap$data$taxon_abund <- calc_taxon_abund(taxmap, "otu_table", cols = row.names(meta_df))

  # calculate wilcoxon rank sum test
  taxmap$data$diff_table <- compare_groups(taxmap, dataset = "taxon_abund",
                                         cols = row.names(meta_df),
                                         groups = meta_df$focused.cluster)

  # adjust p-values
  taxmap <- mutate_obs(taxmap, "diff_table",
                  wilcox_p_value = p.adjust(wilcox_p_value, method = "fdr"))

  # adjust value to zero if not significant p-value
  taxmap$data$diff_table$log2_median_ratio[taxmap$data$diff_table$wilcox_p_value > 0.05] <- 0
  
  # save significant names and values to csv
  save_diff_abund_tax(taxmap, paste("../data/metacoder_differential_abundance/", save_name, ".csv", sep=""))

  return(taxmap)
}
```

```{r}
WRStaxmap_stable_vs_recovering <- wilcoxon_rank_sum_cluster(ps_stable_vs_recovering, "stable_vs_recovering")
WRStaxmap_stable_vs_recovering
```

```{r warning=FALSE}
WRStaxmap_perturbedM_vs_perturbedG <- wilcoxon_rank_sum_focused_cluster(ps_perturbedM_vs_perturbedG, "perturbedM_vs_perturbedG")
WRStaxmap_recovering <- wilcoxon_rank_sum_focused_cluster(ps_recovering, "recovering")
WRStaxmap_recoveringC_vs_recoveringG <- wilcoxon_rank_sum_focused_cluster(ps_recoveringC_vs_recoveringG, "recoveringC_vs_recoveringG")
WRStaxmap_recoveringC_vs_recoveringM <- wilcoxon_rank_sum_focused_cluster(ps_recoveringC_vs_recoveringM, "recoveringC_vs_recoveringM")
WRStaxmap_recoveringG_vs_recoveringM <- wilcoxon_rank_sum_focused_cluster(ps_recoveringG_vs_recoveringM, "recoveringG_vs_recoveringM")
WRStaxmap_stable <- wilcoxon_rank_sum_focused_cluster(ps_stable, "stable")
WRStaxmap_stableC_vs_stableG <- wilcoxon_rank_sum_focused_cluster(ps_stableC_vs_stableG, "stableC_vs_stableG")
WRStaxmap_stableC_vs_stableM <- wilcoxon_rank_sum_focused_cluster(ps_stableC_vs_stableM, "stableC_vs_stableM")
WRStaxmap_stableG_vs_stableM <- wilcoxon_rank_sum_focused_cluster(ps_stableG_vs_stableM, "stableG_vs_stableM")
```

```{r warning=FALSE}
# with clr transform
WRStaxmap_clr_stable_vs_recovering <- wilcoxon_rank_sum_cluster(microbiome::transform(ps_stable_vs_recovering, "clr"), "stable_vs_recovering_clr")
WRStaxmap_clr_perturbedM_vs_perturbedG <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_perturbedM_vs_perturbedG, "clr"), "perturbedM_vs_perturbedG_clr")
WRStaxmap_clr_recovering <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_recovering, "clr"), "recovering_clr")
WRStaxmap_clr_recoveringC_vs_recoveringG <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_recoveringC_vs_recoveringG, "clr"), "recoveringC_vs_recoveringG_clr")
WRStaxmap_clr_recoveringC_vs_recoveringM <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_recoveringC_vs_recoveringM, "clr"), "recoveringC_vs_recoveringM_clr")
WRStaxmap_clr_recoveringG_vs_recoveringM <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_recoveringG_vs_recoveringM, "clr"), "recoveringG_vs_recoveringM_clr")
WRStaxmap_clr_stable <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_stable, "clr"), "stable_clr")
WRStaxmap_clr_stableC_vs_stableG <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_stableC_vs_stableG, "clr"), "stableC_vs_stableG_clr")
WRStaxmap_clr_stableC_vs_stableM <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_stableC_vs_stableM, "clr"), "stableC_vs_stableM_clr")
WRStaxmap_clr_stableG_vs_stableM <- wilcoxon_rank_sum_focused_cluster(microbiome::transform(ps_stableG_vs_stableM, "clr"), "stableG_vs_stableM_clr")
```

# Visualize Results in Heat Trees

Show counts-based and clr-counts-based

## All

```{r}
taxmap %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of rows belonging to taxa (and subtaxa) 
            node_color = n_obs, # number of rows belonging to taxa (and subtaxa) 
            node_color_axis_label = "OTU Counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "All",
            output_file = "../img/metacoder_differential_abundance/heat_tree_all.png")
```

```{r}
taxmap_clr %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of rows belonging to taxa (and subtaxa) 
            node_color = n_obs, # number of rows belonging to taxa (and subtaxa) 
            node_color_axis_label = "OTU Counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "All CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_all_clr.png")
```

## Pairwise -- Stable vs. Recovering

```{r}
(WRStaxmap_stable_vs_recovering)
```

```{r}
WRStaxmap_stable_vs_recovering %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Stable vs. Recovering Samples",
            output_file = "../img/metacoder_differential_abundance/heat_tree_stable_vs_recovering.png")
```

```{r}
WRStaxmap_clr_stable_vs_recovering %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Stable vs. Recovering Samples CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_stable_vs_recovering_clr.png")
```

## Pairwise -- Perturbed M vs. Perturbed G

```{r}
WRStaxmap_perturbedM_vs_perturbedG %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Perturbed M vs. Perturbed G Samples",
            output_file = "../img/metacoder_differential_abundance/heat_tree_perturbedG_vs_perturbedM.png")
```

```{r}
WRStaxmap_clr_perturbedM_vs_perturbedG %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Perturbed M vs. Perturbed G Samples CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_perturbedG_vs_perturbedM_clr.png")
```

## Heat Trees -- Recovering

```{r}
WRStaxmap_recoveringC_vs_recoveringM %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Recovering C vs. Recovering M Samples",
            output_file = "../img/metacoder_differential_abundance/heat_tree_recoveringC_vs_recoveringM.png")
```

```{r}
WRStaxmap_clr_recoveringC_vs_recoveringM %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Recovering C vs. Recovering M Samples CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_recoveringC_vs_recoveringM_clr.png")
```

```{r}
WRStaxmap_recoveringC_vs_recoveringG %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Recovering C vs. Recovering G Samples",
            output_file = "../img/metacoder_differential_abundance/heat_tree_recoveringC_vs_recoveringG.png")
```

```{r}
WRStaxmap_clr_recoveringC_vs_recoveringG %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Recovering C vs. Recovering G Samples CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_recoveringC_vs_recoveringG_clr.png")
```

```{r}
WRStaxmap_recoveringG_vs_recoveringM %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Recovering G vs. Recovering M Samples",
            output_file = "../img/metacoder_differential_abundance/heat_tree_recoveringG_vs_recoveringM.png")
```

```{r}
WRStaxmap_clr_recoveringG_vs_recoveringM %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Recovering G vs. Recovering M Samples CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_recoveringG_vs_recoveringM_clr.png")
```

## Heat Trees -- Stable

```{r}
WRStaxmap_stableC_vs_stableM %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Stable C vs. Stable M Samples",
            output_file = "../img/metacoder_differential_abundance/heat_tree_stableC_vs_stableM.png")
```

```{r}
WRStaxmap_clr_stableC_vs_stableM %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Stable C vs. Stable M Samples CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_stableC_vs_stableM_clr.png")
```

```{r}
WRStaxmap_stableC_vs_stableG %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Stable C vs. Stable G Samples",
            output_file = "../img/metacoder_differential_abundance/heat_tree_stableC_vs_stableG.png")
```

```{r}
WRStaxmap_clr_stableC_vs_stableG %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Stable C vs. Stable G Samples CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_stableC_vs_stableG_clr.png")
```

```{r}
WRStaxmap_stableG_vs_stableM %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Stable G vs. Stable M Samples",
            output_file = "../img/metacoder_differential_abundance/heat_tree_stableG_vs_stableM.png")
```

```{r}
WRStaxmap_clr_stableG_vs_stableM %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree(node_size = n_obs, # number of OTUs
            node_color = log2_median_ratio, # difference between groups
            node_color_interval = c(-10, 10), # symmetric interval
            node_color_range = c("cyan", "gray", "magenta"), # diverging colors
            node_size_axis_label = "OTU count",
            node_color_axis_label = "Log 2 ratio of median counts",
            node_label = taxon_names,
            initial_layout = "reingold-tilford",
            layout = "davidson-harel",
            title = "Stable G vs. Stable M Samples CLR",
            output_file = "../img/metacoder_differential_abundance/heat_tree_stableG_vs_stableM_clr.png")
```

## Heat Matrices

```{r}
theme_set(theme(
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank()))
```

```{r}
WRStaxmap_recovering %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree_matrix(data = "diff_table",
                   node_label = taxon_names,
                   node_size = n_obs, # number of OTUs
                   node_color = log2_median_ratio, # difference between groups
                   node_color_trans = "linear",
                   node_color_interval = c(-3, 3), # symmetric interval
                   edge_color_interval = c(-3, 3), # symmetric interval
                   node_color_range = diverging_palette(), # diverging colors
                   node_size_axis_label = "OTU count",
                   node_color_axis_label = "Log 2 ratio of median counts",
                   layout = "da", initial_layout = "re",
                   output_file = "../img/metacoder_differential_abundance/heat_tree_matrix_recovering.png"
                  )
```

```{r}
WRStaxmap_clr_recovering %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree_matrix(data = "diff_table",
                   node_label = taxon_names,
                   node_size = n_obs, # number of OTUs
                   node_color = log2_median_ratio, # difference between groups
                   node_color_trans = "linear",
                   node_color_interval = c(-3, 3), # symmetric interval
                   edge_color_interval = c(-3, 3), # symmetric interval
                   node_color_range = diverging_palette(), # diverging colors
                   node_size_axis_label = "OTU count",
                   node_color_axis_label = "Log 2 ratio of median counts",
                   layout = "da", initial_layout = "re",
                   output_file = "../img/metacoder_differential_abundance/heat_tree_matrix_recovering_clr.png"
                  )
```

```{r}
WRStaxmap_stable %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree_matrix(data = "diff_table",
                   node_label = taxon_names,
                   node_size = n_obs, # number of OTUs
                   node_color = log2_median_ratio, # difference between groups
                   node_color_trans = "linear",
                   node_color_interval = c(-3, 3), # symmetric interval
                   edge_color_interval = c(-3, 3), # symmetric interval
                   node_color_range = diverging_palette(), # diverging colors
                   node_size_axis_label = "OTU count",
                   node_color_axis_label = "Log 2 ratio of median counts",
                   layout = "da", initial_layout = "re",
                   output_file = "../img/metacoder_differential_abundance/heat_tree_matrix_stable.png"
                  )
```

```{r}
WRStaxmap_clr_stable %>%
  filter_taxa(taxon_names %in% c("bacteria"), subtaxa = TRUE) %>%
  heat_tree_matrix(data = "diff_table",
                   node_label = taxon_names,
                   node_size = n_obs, # number of OTUs
                   node_color = log2_median_ratio, # difference between groups
                   node_color_trans = "linear",
                   node_color_interval = c(-3, 3), # symmetric interval
                   edge_color_interval = c(-3, 3), # symmetric interval
                   node_color_range = diverging_palette(), # diverging colors
                   node_size_axis_label = "OTU count",
                   node_color_axis_label = "Log 2 ratio of median counts",
                   layout = "da", initial_layout = "re",
                   output_file = "../img/metacoder_differential_abundance/heat_tree_matrix_stable_clr.png"
                  )
```
