---
title: "differential abundance wilcoxon sum rank"
output: html_notebook
---

Goal: since there is a significant difference in the composition of clusters, find out who is driving the difference

Groups
* "stable" vs. "recovering"
* "perturbed M" vs. "perturbed G"

* "recovering"
* "recovering C" vs. "recovering G"
* "recovering C" vs. "recovering M"
* "recovering G" vs. "recovering M"

* "stable"
* "stable C" vs. "stable G" 
* "stable C" vs. "stable M"
* "stable G" vs. "stable M"

Methods:
* Wilcox Sum Rank Test
* ALDEx2
* ANCOM-BC

Levels
* Phylum
* Class
* Order
* Family
* Genus
* Species

# Install Packages

```{r}
# .cran_packages <- c("tidyverse", "cowplot", "picante", "vegan", "HMP", "dendextend", "rms", "devtools")
# .inst <- .cran_packages %in% installed.packages()
# if(any(!.inst)) {
#   install.packages(.cran_packages[!.inst])
# }
```

```{r}
# .bioc_packages <- c("phyloseq", "Biostrings", "DESeq2", "microbiome", "metagenomeSeq", "ALDEx2")
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(.bioc_packages, version = "3.14")
```

# Loading Required Packages

```{r}
library("ggplot2"); packageVersion("ggplot2") # plotting
```
```{r}
library("tidyverse"); packageVersion("tidyverse") # util              
```
```{r}
library("dplyr"); packageVersion("dplyr") # util
```
```{r}
library("phyloseq"); packageVersion("phyloseq") # data object                 
```
```{r}
library("ape"); packageVersion("ape") # taxa string manipulation
```
```{r}
library(microbiome); packageVersion("microbiome") # clr transformation
```
```{r}
library(dendextend); packageVersion("dendextend") # extend customization of dendograms
```
```{r}
library(vegan); packageVersion("vegan") # bray-curtis and other metrics
```

```{r}
# Manipulation and visualization of taxonomic data, particularly those from amplicon metagenomics research. (Foster, Sharpton, and Grünwald 2017)
library(metacoder); packageVersion("metacoder")
```
```{r}
# defines taxonomic classes and functions to manipulate them. The goal is to use these classes as low level fundamental taxonomic classes that other R packages can build on and use. This is used by metacoder. 
library(taxa); packageVersion("taxa")
```
```{r}
library(ALDEx2); packageVersion("ALDEx2") # aldex2 test (compositional differential abundance)
```



# Initialization

```{r}
# ggplot2 package theme set
theme_set(theme_bw())

# ensure randomization is same
set.seed(42)
```

# Load Phyloseq Data Object
```{r}
(ps <- readRDS("../data/processed/phyloseq.rds"))
# convert to taxmap format 
taxmap <- parse_phyloseq(ps)

# Get Relative Abundance transform values
ps_rel_abund = phyloseq::transform_sample_counts(ps, function(x){x / sum(x)})

# Get CLR transform values
ps_clr <- microbiome::transform(ps, "clr")
taxmap_clr <- parse_phyloseq(ps_clr)
```

# Make Subset Cluster Group Tables

* "stable" vs. "recovering"
```{r}
(ps_stable_vs_recovering <- phyloseq::subset_samples(ps, cluster == "stable" | cluster == "recovering"))
(ps_stable_vs_recovering <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stable_vs_recovering) > 0, ps_stable_vs_recovering))
```

* "perturbed M" vs. "perturbed G"
```{r}
(ps_perturbedM_vs_perturbedG <- phyloseq::subset_samples(ps, focused.cluster == "perturbed M" | focused.cluster == "perturbed G"))
(ps_perturbedM_vs_perturbedG <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_perturbedM_vs_perturbedG) > 0, ps_perturbedM_vs_perturbedG))
```
* "recovering"
* "recovering C" vs. "recovering G"
* "recovering C" vs. "recovering M"
* "recovering G" vs. "recovering M"
```{r}
(ps_recovering <- phyloseq::subset_samples(ps, cluster == "recovering"))
(ps_recovering <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_recovering) > 0, ps_recovering))

(ps_recoveringC_vs_recoveringG <- phyloseq::subset_samples(ps, focused.cluster == "recovering C" | focused.cluster == "recovering G"))
(ps_recoveringC_vs_recoveringG <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_recoveringC_vs_recoveringG) > 0, ps_recoveringC_vs_recoveringG))

(ps_recoveringC_vs_recoveringM <- phyloseq::subset_samples(ps, focused.cluster == "recovering C" | focused.cluster == "recovering M"))
(ps_recoveringC_vs_recoveringM <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_recoveringC_vs_recoveringM) > 0, ps_recoveringC_vs_recoveringM))

(ps_recoveringG_vs_recoveringM <- phyloseq::subset_samples(ps, focused.cluster == "recovering G" | focused.cluster == "recovering M"))
(ps_recoveringG_vs_recoveringM <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_recoveringG_vs_recoveringM) > 0, ps_recoveringG_vs_recoveringM))
```
* "stable"
* "stable C" vs. "stable G", 
* "stable C" vs. "stable M"
* "stable G" vs. "stable M"
```{r}
(ps_stable <- phyloseq::subset_samples(ps, cluster == "stable"))
(ps_stable <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stable) > 0, ps_stable))

(ps_stableC_vs_stableG <- phyloseq::subset_samples(ps, focused.cluster == "stable C" | focused.cluster == "stable G"))
(ps_stableC_vs_stableG <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stableC_vs_stableG) > 0, ps_stableC_vs_stableG))

(ps_stableC_vs_stableM <- phyloseq::subset_samples(ps, focused.cluster == "stable C" | focused.cluster == "stable M"))
(ps_stableC_vs_stableM <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stableC_vs_stableM) > 0, ps_stableC_vs_stableM))

(ps_stableG_vs_stableM <- phyloseq::subset_samples(ps, focused.cluster == "stable G" | focused.cluster == "stable M"))
(ps_stableG_vs_stableM <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps_stableG_vs_stableM) > 0, ps_stableG_vs_stableM))
```

# ALDEx2 Test

```{r}
aldex2_cluster <- function(ps, save_name) {
  #Run ALDEx2
  aldex2_da <- ALDEx2::aldex(data.frame(phyloseq::otu_table(ps)), phyloseq::sample_data(ps)$cluster, test="t", effect = TRUE, denom="iqlr")
  
  #Plot effect sizes
  png(file=paste("../img/aldex2_differential_abundance/", save_name, ".png", sep=""))
  par(mfrow=c(1,2))
  ALDEx2::aldex.plot(aldex2_da, type="MA", test="wilcox")
  ALDEx2::aldex.plot(aldex2_da, type="MW", test="wilcox")
  dev.off()
  
  #Clean up presentation
  sig_aldex2 <- aldex2_da %>%
    rownames_to_column(var = "OTU") %>%
    filter(wi.eBH < 0.05) %>%
    arrange(effect, wi.eBH) %>%
    dplyr::select(OTU, diff.btw, diff.win, effect, wi.ep, wi.eBH)
  
  # save significant names and values to csv
  sig_aldex2 <- merge(x=sig_aldex2, y=phyloseq::tax_table(ps), by.x="OTU", by.y=0, all.x=TRUE)
  write.csv(sig_aldex2, paste("../data/aldex2_differential_abundance/", save_name, ".csv", sep=""))
  
  return(sig_aldex2)
}


aldex2_focused_cluster <- function(ps, save_name) {
   #Run ALDEx2
  aldex2_da <- ALDEx2::aldex(data.frame(phyloseq::otu_table(ps)), phyloseq::sample_data(ps)$focused.cluster, test="t", effect = TRUE, denom="iqlr")
  
  #Plot effect sizes
  png(file=paste("../img/aldex2_differential_abundance/", save_name, ".png", sep=""))
  par(mfrow=c(1,2))
  ALDEx2::aldex.plot(aldex2_da, type="MA", test="wilcox")
  ALDEx2::aldex.plot(aldex2_da, type="MW", test="wilcox")
  dev.off()
  
  #Clean up presentation
  sig_aldex2 <- aldex2_da %>%
    rownames_to_column(var = "OTU") %>%
    filter(wi.eBH < 0.05) %>%
    arrange(effect, wi.eBH) %>%
    dplyr::select(OTU, diff.btw, diff.win, effect, wi.ep, wi.eBH)
  
  # save significant names and values to csv
  sig_aldex2 <- merge(x=sig_aldex2, y=phyloseq::tax_table(ps), by.x="OTU", by.y=0, all.x=TRUE)
  write.csv(sig_aldex2, paste("../data/aldex2_differential_abundance/", save_name, ".csv", sep=""))
  
  return(sig_aldex2)
}
```

```{r}
aldex2_stable_vs_recovering <- aldex2_cluster(ps_stable_vs_recovering, "stable_vs_recovering")
aldex2_stable_vs_recovering
```

```{r warning=FALSE}
aldex2_perturbedM_vs_perturbedG <- aldex2_focused_cluster(ps_perturbedM_vs_perturbedG, "perturbedM_vs_perturbedG")
aldex2_recoveringC_vs_recoveringG <- aldex2_focused_cluster(ps_recoveringC_vs_recoveringG, "recoveringC_vs_recoveringG")
aldex2_recoveringC_vs_recoveringM <- aldex2_focused_cluster(ps_recoveringC_vs_recoveringM, "recoveringC_vs_recoveringM")
aldex2_recoveringG_vs_recoveringM <- aldex2_focused_cluster(ps_recoveringG_vs_recoveringM, "recoveringG_vs_recoveringM")
aldex2_stableC_vs_stableG <- aldex2_focused_cluster(ps_stableC_vs_stableG, "stableC_vs_stableG")
aldex2_stableC_vs_stableM <- aldex2_focused_cluster(ps_stableC_vs_stableM, "stableC_vs_stableM")
aldex2_stableG_vs_stableM <- aldex2_focused_cluster(ps_stableG_vs_stableM, "stableG_vs_stableM")
```

